\setstretch{1.5}
\setlength{\parskip}{0.5\baselineskip}

This chapter describes the experimental methodology of each phase.
For every phase, the structure follows a consistent template: hypothesis,
method (step-by-step), variables, output files, and the research question
answered.
Section~\ref{sec:design} explains the overarching design principles that
govern how the phases relate to each other.
Sections~\ref{sec:phase1} through~\ref{sec:phase6} describe each phase in
full. Table~\ref{tab:phase_summary} at the end of this chapter provides a
consolidated reference.

% ============================================================
\section{Experimental Design Principles}
\label{sec:design}
% ============================================================

\subsection{The Isolation Principle}
\label{subsec:isolation}

The defining design choice of this study is that Phases~3, 4, and~5 are
\emph{isolated} experiments. Each phase changes exactly one variable relative
to Phase~2: the type of knowledge injected into the system prompt or applied
as a post-processing step. Every other factor --- the model, the dataset,
the sample set, the generation parameters, the evaluation protocol, the
analysis code --- is held fixed. This design allows any observed CER change
to be attributed unambiguously to the knowledge injection and not to a
confound.

Comparing isolated phases against Phase~2 (zero-shot) rather than against
Phase~1 (OCR baseline) is a deliberate choice. The question of interest is
not whether the LLM improves on raw OCR (Phase~2 answers that), but whether
injecting a specific type of knowledge into an already-running LLM corrector
improves its performance. Comparing to Phase~1 would confound OCR quality
with LLM correction quality.

\subsection{Phase~2 as the Hub Baseline}
\label{subsec:hub}

Phase~2 (zero-shot LLM correction) serves as the comparison baseline for all
subsequent phases. A phase that does not improve on Phase~2 is a negative
result with respect to that knowledge type: the injected knowledge does not
help, and may actively hurt, relative to the model's own language
understanding. Negative results are scientifically valuable because they
rule out a hypothesis and constrain theories of what makes LLM correction
succeed. All five strategies are therefore compared to the same zero-shot
baseline, enabling a consistent ranking.

\subsection{Reproducibility Measures}
\label{subsec:reproducibility}

All experiments use:
\begin{itemize}
  \item The same nine evaluation datasets: eight PATS-A01 fonts (415 validation
        samples each, aligned split) plus KHATT validation.
  \item The same model: \texttt{Qwen/Qwen3-4B-Instruct-2507}.
  \item The same generation parameters: temperature 0.1,
        \texttt{max\_new\_tokens} 1024, \texttt{enable\_thinking=False}.
  \item The same evaluation protocol: diacritic-stripped CER/WER, per-sample
        then aggregated. PATS-A01 macro-averages (non-weighted, each font
        contributes equally) are reported alongside per-font results.
  \item Fixed random seeds where any sampling is involved (seed 42 throughout).
\end{itemize}
Each phase stores its inference input as a JSONL file with embedded prompts,
ensuring that the exact prompts sent to the model are permanently recorded
and can be inspected or re-run without requiring access to the original data
files.

\subsection{Why Not Test All Combinations First}
\label{subsec:why_not_all}

With three distinct prompt-level knowledge sources (confusion matrix,
self-reflective insights, and their combination), plus CAMeL post-processing
as an optional layer, the combination space is $2^3 = 8$ prompt subsets plus
a post-processing overlay. Running all 8 combinations across nine datasets
as the primary experimental strategy would require approximately 33,000 model
inference calls --- well beyond the available compute budget for a
preliminary sweep --- and, more importantly, would make results difficult to
interpret: a combination that degrades performance could not be decomposed
into its contributing factors without the isolated phase results.

The hierarchical strategy (isolated first, then selected combinations in
Phase~6) extracts the maximum insight per inference call. Isolated phases
diagnose each component independently. Phase~6 then tests four pre-selected
combinations chosen based on the isolated results: the best-performing
isolated strategy, the second-best, their combination, and the combination
augmented with CAMeL post-processing. This focused design answers the
combination question with a fraction of the inference cost of a full grid
search.

% ============================================================
\section{Phase 1 --- Baseline and Error Taxonomy}
\label{sec:phase1}
% ============================================================

\subsection{Hypothesis}
Qaari produces systematic, categorisable errors on Arabic typewritten text.
The error profile is dominated by a small number of recurring character
confusion types that can be quantified and used to guide subsequent phases.
The nature and frequency of errors differs substantially across fonts, making
multi-font evaluation necessary to capture the full range of correction
challenges.

\subsection{Method}
\label{subsec:phase1_method}

\begin{enumerate}
  \item Load all OCR predictions and ground-truth pairs for each of the nine
        evaluation datasets from the DataLoader. Both training and validation
        splits are processed; only validation-split metrics are reported in
        Chapter~\ref{ch_results}, but training-split confusion matrices and
        error statistics inform Phases~3 and~4.

  \item Compute per-sample CER and WER using the normalised metric
        (diacritics stripped via Unicode category filtering).
        Aggregate statistics per dataset: mean, standard deviation, median,
        95th percentile (P95). The P95 metric identifies the severity of
        the worst-performing tail of samples.

  \item Build a character-level confusion matrix.
        For each aligned (ground-truth character, OCR character) pair in the
        minimum-edit alignment, increment the count for that pair.
        Normalise by ground-truth character frequency to obtain conditional
        confusion probabilities. The confusion matrix is stored as a
        square matrix indexed by Unicode character code point and is used
        directly by Phase~3's \texttt{ConfusionMatrixLoader}.

  \item Classify each substitution, insertion, and deletion into one of eight
        taxonomic categories using the heuristic rules defined in
        Chapter~\ref{ch_system}:
        \begin{enumerate}
          \item Dot confusion (ba/ta/tha/ya family; fa/qaf; nun/waw variants)
          \item Hamza and alef variants (hamza placement, alef/alef-madda/alef-wasla)
          \item Taa marbuta / ha confusion
          \item Alef maqsura / ya confusion
          \item Similar-shape pairs (ra/zay, dal/dhal, sad/dad, etc.)
          \item Word merge (two words recognised as one)
          \item Word split (one word recognised as two)
          \item Other substitution, insertion, or deletion
        \end{enumerate}

  \item Apply CAMeL Tools morphological analysis to each unique OCR token.
        Classify tokens as \emph{non-words} (zero valid analyses returned by
        the MSA morphological analyser) or \emph{valid-but-wrong} (at least
        one valid analysis, but contextually incorrect relative to ground
        truth). This distinction is critical for Phase~5: the CAMeL
        post-processing revert strategy can only intervene on non-word errors.

  \item Cross-reference the confusion matrix with Phase~2 training-split
        failures: retain only confusion pairs that appear in samples where
        the zero-shot LLM failed to correct the error. This filtered confusion
        list is used by Phase~3 and represents actionable patterns --- ones
        the model is known not to handle by default.

  \item Write per-dataset reports including confusion matrix JSON, error
        taxonomy JSON, morphological analysis JSON, and a human-readable
        Markdown summary.
\end{enumerate}

\subsection{Output Files}
\label{subsec:phase1_outputs}
\begin{itemize}
  \item \texttt{results/phase1/\{dataset\}/metrics.json} --- CER/WER statistics
        (mean, std, median, p95) with and without diacritics
  \item \texttt{results/phase1/\{dataset\}/confusion\_matrix.json} --- character-level
        confusion counts with top-20 list and conditional confusion probabilities
  \item \texttt{results/phase1/\{dataset\}/error\_taxonomy.json} --- error counts
        and percentages by eight-category taxonomy with representative examples
  \item \texttt{results/phase1/\{dataset\}/morphological\_analysis.json} --- word
        validity statistics (non-word / valid-but-wrong split)
  \item \texttt{results/phase1/\{dataset\}/error\_examples.json} --- up to 5 sample
        OCR/GT pairs per error category
  \item \texttt{results/phase1/\{dataset\}/report.md} --- human-readable summary
\end{itemize}

\noindent
\textbf{Research question answered:} RQ1.

% ============================================================
\section{Phase 2 --- Zero-Shot LLM Correction}
\label{sec:phase2}
% ============================================================

\subsection{Hypothesis}
A zero-shot instruction-tuned LLM can improve Arabic OCR output without
task-specific guidance, by exploiting its pre-trained Arabic language knowledge
to recognise and correct implausible character sequences. However, the
effectiveness of this correction is expected to be font-dependent: fonts with
higher baseline CER present more recognisable errors, while fonts with lower
CER present fewer errors per sample and risk over-correction.

\subsection{Method}
\label{subsec:phase2_method}

\begin{enumerate}
  \item For each sample in each of the nine evaluation datasets, construct the
        conservative zero-shot prompt using
        \texttt{PromptBuilder.build\_zero\_shot(ocr\_text, version="v2")}.
        The system message informs the model that the text is OCR output that
        may already be correct, instructs it to return the text unchanged if
        no clear error is found, and prohibits it from changing what is
        already correct.

  \item Write all prompts to
        \texttt{results/phase2/inference\_input.jsonl}, one JSON object per
        sample, containing fields: \texttt{sample\_id}, \texttt{dataset},
        \texttt{ocr\_text}, \texttt{gt\_text}, \texttt{prompt},
        \texttt{prompt\_version} (= \texttt{p2v2}).

  \item Run inference on Kaggle (GPU) using \texttt{scripts/infer.py}.
        For each sample, the model generates the corrected text.
        Results are written to \texttt{results/phase2/corrections.jsonl}.
        The inference script implements a resume mechanism: any
        \texttt{sample\_id} already present in the output file is skipped,
        enabling restartable inference without duplicating work.

  \item Compute post-correction CER and WER for each sample.
        Compare to Phase~1 OCR baseline.
        Compute error changes per sample: fixed errors (present in OCR,
        absent from correction), introduced errors (absent from OCR,
        present in correction), and their ratio (introduction rate).

  \item Run the same Phase~2 pipeline on the \emph{training} split of each
        PATS-A01 font. These training-split corrections are not reported as
        results but serve as the input to Phase~4's self-reflective analysis.

  \item Write comparison reports to
        \texttt{results/phase2/\{dataset\}/comparison\_vs\_phase1.json}.
\end{enumerate}

\subsection{Prompt Design Rationale}
\label{subsec:phase2_rationale}

Two prompt versions were developed and piloted:
\begin{itemize}
  \item \textbf{v1 (aggressive)}: instructs the model to correct OCR errors
        without qualification. Piloting on Akhbar (low baseline CER) showed
        that v1 increased CER relative to baseline: the model rewrote
        correctly recognised text. This aggressive framing was discarded.
  \item \textbf{v2 (conservative)}: explicitly tells the model that the text
        \emph{may already be correct}, that it should return it unchanged if
        no clear error is visible, and that it must not change what is
        correct. This formulation keeps the model in a low-intervention mode
        on clean text while still enabling correction where errors are obvious.
\end{itemize}
Version v2 is used in all phases reported in this thesis. The trade-off is
that v2 may miss subtle errors that v1 would catch, but the reduction in
over-correction is empirically justified.

\subsection{Output Files}
\begin{itemize}
  \item \texttt{results/phase2/\{dataset\}/metrics.json}
  \item \texttt{results/phase2/\{dataset\}/comparison\_vs\_phase1.json}
  \item \texttt{results/phase2/\{dataset\}/error\_changes.json}
\end{itemize}

\noindent
\textbf{Research question answered:} RQ2.
\textbf{Role in the study:} Phase~2 output serves as the comparison baseline
for all subsequent phases and as the input data source for Phase~4's
training-split analysis.

% ============================================================
\section{Phase 3 --- OCR-Aware Prompting}
\label{sec:phase3}
% ============================================================

\subsection{Hypothesis}
Telling the LLM which specific character confusions Qaari makes on a given
font --- filtered to confusions that appear in samples the model itself
failed to correct at the zero-shot level --- will direct its attention to
the positions most likely to contain unresolved errors and improve correction
quality for those specific confusion types.

\subsection{Method}
\label{subsec:phase3_method}

\begin{enumerate}
  \item Load the Phase~1 training-split confusion matrix from
        \texttt{results/phase1/\{dataset\}-train/confusion\_matrix.json}
        using \texttt{ConfusionMatrixLoader}. Each font uses its own
        font-specific confusion matrix derived from its training split,
        ensuring the confusion context is tailored to the actual error
        distribution of that typeface.

  \item Cross-reference the confusion matrix entries with the LLM failure
        cases from Phase~2 training-split analysis. A confusion pair
        (gt\_char, ocr\_char) is retained in the injection list only if it
        appears in samples where the zero-shot model failed to produce a
        correction closer to ground truth than the OCR input. This
        \emph{failure-filtered} list focuses the model's attention on
        confusions it does not resolve by default.

  \item Extract the top-10 confusion pairs by substitution count from the
        filtered list.

  \item Format the confusion pairs as an Arabic-language bullet list using
        \texttt{ConfusionMatrixLoader.format\_for\_prompt()}.
        Each bullet item reads (translated): ``This system confuses
        [correct character] with [OCR character] in output.''

  \item Build the OCR-aware prompt using
        \texttt{PromptBuilder.build\_ocr\_aware(ocr\_text, confusion\_context)}.
        The system message extends the zero-shot v2 message with the
        confusion list in a \texttt{<confusion\_patterns>} XML block,
        instructing the model to pay particular attention to these pairs
        while observing the same conservative correction rules as Phase~2.

  \item Export, infer (Kaggle GPU), and analyse following the same three-stage
        pipeline as Phase~2. The prompt version tag is \texttt{p3v1}.
\end{enumerate}

\subsection{Variables}
\begin{itemize}
  \item \textbf{Changed from Phase~2:} Failure-filtered confusion list
        injected into system prompt.
  \item \textbf{Fixed:} Model, dataset, generation parameters, evaluation
        protocol, conservative correction instruction.
  \item \textbf{Number of confusion pairs injected:} 10 (configured as
        \texttt{phase3.top\_n} in \texttt{configs/config.yaml}).
\end{itemize}

\subsection{Design Rationale}
\label{subsec:phase3_rationale}

The failure-filtering step is the key enhancement in Phase~3. An unfiltered
confusion matrix lists all character confusions observed in the training data,
including confusions that the zero-shot model already handles correctly. Such
confusions carry no additional information for the model. By retaining only
confusions from samples where the model failed, the injection is narrowed to
genuine blind spots. This makes the injected list shorter and more targeted
than a naive top-10 from the raw confusion matrix.

The font-specificity of the confusion matrix is equally important. A single
confusion list shared across all fonts would reflect the aggregate Qaari error
profile, which may not match the dominant errors on any individual font. Using
per-font confusion matrices ensures that each font receives a confusion context
that matches its specific OCR error distribution.

\subsection{Output Files}
\begin{itemize}
  \item \texttt{results/phase3/\{dataset\}/metrics.json}
  \item \texttt{results/phase3/\{dataset\}/comparison\_vs\_phase2.json}
  \item \texttt{results/phase3/\{dataset\}/confusion\_impact.json} ---
        per-confusion-pair analysis of correction improvement on samples
        containing that confusion type
\end{itemize}

\noindent
\textbf{Research question answered:} RQ3.

% ============================================================
\section{Phase 4 --- Self-Reflective Prompting}
\label{sec:phase4}
% ============================================================

\subsection{Hypothesis}
Feeding the model statistical summaries of its own systematic errors ---
derived from its zero-shot corrections on the training split where ground
truth is available --- will cause it to apply targeted corrections more
carefully and reduce its error introduction rate on the held-out validation
split. Additionally, providing concrete word-level OCR error pairs from the
training data gives the model specific exemplars of the correction patterns
it should apply, beyond what aggregate statistics alone can convey.

\subsection{Three-Stage Method}
\label{subsec:phase4_method}

Phase~4 is itself a three-stage process that must be executed in order.
The first stage uses training-split data; the second and third stages use
validation-split data.

\subsubsection{Stage~1: Training-Split Analysis}

\begin{enumerate}
  \item Load Phase~2 corrections on the \emph{training} split for each
        PATS-A01 font (e.g.,
        \texttt{results/phase2-training/\{font\}-train/corrections.jsonl}).
        These corrections exist because Phase~2 is run on both splits;
        the training-split corrections are not used as results but solely
        as analysis data.

  \item For each sample, compare the Phase~2 LLM corrections to ground truth
        using \texttt{LLMErrorAnalyzer.analyse\_sample()}.
        For every aligned character pair between LLM output and ground truth:
        \begin{itemize}
          \item If the OCR had an error at this position and the LLM fixed it:
                increment \texttt{fixed} for that error type.
          \item If the OCR was correct at this position and the LLM changed it
                incorrectly: increment \texttt{introduced} for that error type.
        \end{itemize}

  \item Aggregate across all training samples to obtain per-error-type
        statistics:
        \begin{itemize}
          \item \texttt{fix\_rate} = \texttt{fixed} / baseline errors of that
                type (fraction of existing errors the LLM corrected).
          \item \texttt{introduction\_rate} = \texttt{introduced} /
                ground-truth tokens of that type (fraction of correct tokens
                incorrectly changed).
        \end{itemize}

  \item Extract word-level failure pairs: (ocr\_word, gt\_word) pairs from
        samples that the LLM either failed to fix or made worse. These
        are stored in \texttt{word\_pairs\_llm\_failures.txt} with separate
        sections for UNFIXED errors (LLM left OCR errors unchanged) and
        INTRODUCED errors (LLM introduced new errors not in OCR).

  \item Classify training samples by correction outcome into:
        correctly improved, made worse, and unchanged. Store classification
        in \texttt{sample\_classification.json}.

  \item Aggregate insights separately for PATS-A01 and KHATT datasets,
        reflecting their substantially different error profiles.
\end{enumerate}

\subsubsection{Insights Derived from Training Data}

Table~\ref{tab:phase4_insights} reports representative statistics extracted
from the PATS-A01 training split (aggregated across fonts).

\begin{table}[ht]
\caption{Phase~4 training-split analysis results (PATS-A01 aggregate).
         Fix rate = fraction of baseline errors of that type
         corrected by the LLM. Introduction rate = fraction of correct
         ground-truth tokens of that type incorrectly changed.
         $\dagger$ = fewer than 10 baseline samples (estimate unreliable).}
\label{tab:phase4_insights}
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lrrrrr}
\toprule
\textbf{Error type} & \textbf{Baseline} & \textbf{Fixed} & \textbf{Introduced}
  & \textbf{Fix rate} & \textbf{Intro.\ rate} \\
\midrule
Insertion       & 1,964 & 1,912 & 356  & 97.35\% & 18.13\% \\
Other subst.    &    49 &     8 &  80  & 16.33\% & 163.27\% \\
Deletion        &    32 &     1 &  68  &  3.12\% & 212.50\% \\
Dot confusion   &    14 &     1 &   2  &  7.14\% &  14.29\% \\
Alef maqsura    & 3$^\dagger$ & 1 & 3 & 33.33\% & 100.00\% \\
Similar shape   & 1$^\dagger$ & 1 & 8 & 100.00\% & 800.00\% \\
Hamza           &     0 &   --- & ---  &    ---  &    --- \\
Taa marbuta     &     0 &   --- & ---  &    ---  &    --- \\
\midrule
\textbf{Overall} & \textbf{2,063} & \textbf{1,924} & \textbf{517}
  & \textbf{93.26\%} & \textbf{25.06\%} \\
\bottomrule
\end{tabular}
\end{table}

The table reveals two important patterns. First, the insertion fix rate of
97.35\% shows that the LLM is highly effective at collapsing the Qaari
repetition artefact: almost all spurious character insertions (which
constitute the majority of all baseline errors) are removed. Second, the
overall introduction rate of 25.06\% --- approximately one in four previously
correct tokens is incorrectly changed by the LLM --- represents a systematic
over-correction bias that affects all error categories. The deletion and
other-substitution introduction rates exceed 100\% because the LLM introduces
more of those error types than existed in the original OCR output.

These patterns motivate the self-reflective prompt: by explicitly informing
the model of its high introduction rate and its weakness on deletion and
substitution errors, the prompt aims to shift the model towards a more
conservative correction mode.

\subsubsection{Stage~2: Export}

\begin{enumerate}
  \item Load the aggregated training-split artifacts:
        \texttt{word\_pairs\_llm\_failures.txt} and
        \texttt{sample\_classification.json} from
        \texttt{results/phase2-training/analysis/}.

  \item Apply the insight filtering logic to derive the self-analysis context:
        \begin{itemize}
          \item Error types with fix rate $> 0.60$: reported as strengths
                (``The system is good at fixing this type of error'').
          \item Error types with fix rate $< 0.40$ and baseline $\geq 10$:
                reported as weaknesses (``The system often fails to fix
                this type of error'').
          \item Error types with introduction rate $> 0.05$ and baseline
                $\geq 10$: reported as over-correction risks (``The system
                tends to incorrectly change correct tokens of this type'').
        \end{itemize}

  \item Convert the filtered statistics to Arabic-language diagnostic sentences
        that describe the model's behaviour patterns and provide concrete
        guidance for what to watch out for. The full diagnostic text is
        provided in Appendix~B.

  \item Select up to 15 word-level error pairs per sample from
        \texttt{word\_pairs\_llm\_failures.txt} to include as concrete
        exemplars in the prompt, showing specific (ocr\_word, gt\_word)
        pairs the model has struggled with.

  \item Build the self-reflective prompt for each validation-split sample
        using
        \texttt{PromptBuilder.build\_self\_reflective(ocr\_text, insights\_context)}.
        The insights are injected in a \texttt{<self\_analysis>} XML block
        following the conservative correction instruction from Phase~2.

  \item Write to \texttt{results/phase4/inference\_input.jsonl}.
        The prompt version tag is \texttt{p4v1}.
\end{enumerate}

\subsubsection{Stage~3: Analyse}

\begin{enumerate}
  \item Run inference on Kaggle GPU using \texttt{scripts/infer.py}.
  \item Load \texttt{results/phase4/corrections.jsonl}.
  \item Compute CER/WER per dataset; compare to Phase~2 zero-shot baseline.
  \item Write reports to \texttt{results/phase4/\{dataset\}/}.
\end{enumerate}

\subsection{Variables}
\begin{itemize}
  \item \textbf{Changed from Phase~2:} Arabic-language diagnostic text
        injected in \texttt{<self\_analysis>} block; concrete word-level
        error pairs from training data appended.
  \item \textbf{Fixed:} Model, dataset, generation parameters, evaluation
        protocol, conservative correction instruction.
  \item \textbf{Data used:} Phase~2 training-split corrections (not
        validation-split; no data leakage).
\end{itemize}

\subsection{No-Leakage Guarantee}

A potential concern with self-reflective prompting is data leakage: if the
training and validation splits are not properly separated, the model might
receive information about the ground truth of its validation-split samples
via the error statistics. This concern is addressed by the aligned split
design: the 85\% training split and 15\% validation split share no samples.
The training-split analysis is performed exclusively on training-split
corrections, and the resulting insights are qualitative statistical summaries
(e.g., ``the model has a 25\% introduction rate'') rather than sample-level
ground-truth information. The word-level error pairs are drawn from training
samples only. There is no mechanism by which Phase~4 could convey
ground-truth information about validation samples to the model.

\subsection{Output Files}
\begin{itemize}
  \item \texttt{results/phase2-training/analysis/word\_pairs\_llm\_failures.txt}
        --- UNFIXED and INTRODUCED sections
  \item \texttt{results/phase2-training/analysis/sample\_classification.json}
  \item \texttt{results/phase4/insights/PATS-A01\_insights.json} ---
        aggregated training-split statistics
  \item \texttt{results/phase4/\{dataset\}/metrics.json}
  \item \texttt{results/phase4/\{dataset\}/comparison\_vs\_phase2.json}
  \item \texttt{results/phase4/\{dataset\}/error\_changes.json}
\end{itemize}

\noindent
\textbf{Research question answered:} RQ4.

% ============================================================
\section{Phase 5 --- CAMeL Morphological Post-Processing}
\label{sec:phase5}
% ============================================================

\subsection{Hypothesis}
Morphological post-processing can detect two classes of LLM output failures
and reverse them. First, cases where the LLM introduces morphologically
invalid Arabic words from valid OCR sources can be reverted to the OCR
original. Second, documented overcorrection patterns identified in the
training split can be reverted regardless of morphological status, targeting
the model's systematic tendency to change specific correct word forms.
Together, these two revert conditions reduce the error introduction rate
without substantially affecting correctly corrected words.

\subsection{Method}
\label{subsec:phase5_method}

Phase~5 does not generate new prompts or run a new inference call from
scratch. It applies the revert strategy as a post-processing step to the
Phase~2 corrections.

\begin{enumerate}
  \item Load Phase~2 corrections from
        \texttt{results/phase2/\{dataset\}/corrections.jsonl}.

  \item Load the known-overcorrection list derived from training-split
        analysis. This list contains (ocr\_word, llm\_word) pairs where
        the LLM changed a correct OCR word to an incorrect form,
        identified by comparing Phase~2 training-split corrections to
        ground truth.

  \item Initialise \texttt{MorphAnalyzer} with the MSA database
        \texttt{calima-msa-r13}. CAMeL Tools requires separate installation:
        \texttt{pip install camel-tools} followed by
        \texttt{camel\_data -i morphology-db-msa-r13}.

  \item For each sample, apply
        \texttt{WordValidator.validate\_correction(ocr\_text, llm\_text,
        known\_overcorrections)}:
        \begin{enumerate}
          \item Tokenise both the OCR text and the LLM correction into
                Arabic word tokens, preserving position indices.
          \item For each word pair at position $i$ where
                (ocr\_word$_i$, llm\_word$_i$) differs:
                \begin{itemize}
                  \item \textbf{Morphological check:} Run
                        \texttt{MorphAnalyzer.analyse(llm\_word$_i$)};
                        if zero analyses: LLM word is invalid. Run
                        \texttt{MorphAnalyzer.analyse(ocr\_word$_i$)};
                        if $\geq 1$ analysis: OCR word is valid.
                        If both hold: revert to \texttt{ocr\_word$_i$}.
                  \item \textbf{Known-overcorrection check:} If
                        (ocr\_word$_i$, llm\_word$_i$) appears in the
                        overcorrection list: revert to \texttt{ocr\_word$_i$}
                        regardless of morphological status.
                  \item \textbf{Otherwise:} Keep the LLM correction.
                \end{itemize}
          \item Record per-sample statistics: words reverted by morphological
                check, words reverted by overcorrection check, words kept.
        \end{enumerate}

  \item Compute post-validation CER/WER on the final text.
        Compare to both Phase~2 (LLM only) and Phase~1 (OCR baseline).

  \item Write validation statistics to
        \texttt{results/phase5/\{dataset\}/validation\_stats.json}.
\end{enumerate}

\subsection{Implementation Note: Position-Indexed Revert}

An important implementation detail is that the revert decision is indexed by
\emph{word position}, not word identity. An early implementation keyed reverts
by word string, which caused incorrect reversions when the same word form
appeared multiple times in a sentence and only some occurrences should be
reverted. The position-indexed implementation ensures that each word is
evaluated independently at its specific position, preventing this
cross-position contamination.

\subsection{Revert Strategy Rationale}
\label{subsec:phase5_rationale}

The morphological revert strategy is deliberately conservative. It can only
revert a word to the OCR original; it cannot introduce new words or produce
word forms not present in either the OCR output or the LLM correction.
The maximum harm it can do is revert a correctly-improved LLM word back to
a wrong OCR word --- but only if the LLM word happened to be morphologically
invalid despite being the correct intended word, which is rare for genuine
corrections of real Arabic. The maximum benefit is preventing a hallucinated,
morphologically invalid LLM word from degrading the text.

The known-overcorrection check adds a second layer of protection targeting
the model's documented systematic biases. Because overcorrection patterns
are mined from training data where ground truth is available, they represent
genuine cases where the model has been shown to make specific incorrect
changes. Reverting these patterns on the validation set is a low-risk
intervention, though it covers only a subset of over-corrections (those that
are \emph{also} frequent enough to appear in the training analysis).

\subsection{Expected Limitations}

The primary limitation of Phase~5 is that the dominant OCR error type ---
valid-but-wrong substitutions (71\% of all errors per the Phase~1 analysis)
--- is invisible to a context-free morphological check. Both the original
OCR word and its correct counterpart are morphologically valid, so the
check cannot distinguish them. This means Phase~5's revert mechanism
operates on the minority (29\%) of errors that produce non-words, and the
overcorrection list covers a further small fraction. The expectation is
therefore that Phase~5 will have a modest impact, and may even slightly
degrade performance if reverts accidentally undo beneficial corrections.

\subsection{Output Files}
\begin{itemize}
  \item \texttt{results/phase5/\{dataset\}/metrics.json}
  \item \texttt{results/phase5/\{dataset\}/comparison\_vs\_phase2.json}
  \item \texttt{results/phase5/\{dataset\}/validation\_stats.json} ---
        per-dataset revert counts, rates, and list of reverted word pairs
\end{itemize}

\noindent
\textbf{Research question answered:} RQ5.

% ============================================================
\section{Phase 6 --- Combinations and Ablation}
\label{sec:phase6}
% ============================================================

\subsection{Purpose}
\label{subsec:phase6_purpose}

Phases~3--5 establish which knowledge types help and which hurt in isolation.
Phase~6 asks two follow-up questions:
(a)~can two or more knowledge types combine synergistically to produce a result
better than either individually?
(b)~in the best combined system, what is each component's quantitative
contribution when removed?

\subsection{Combinations Tested}
\label{subsec:phase6_combos}

Four combinations are defined and tested:

\begin{enumerate}
  \item \textbf{\texttt{conf\_only}}: Phase~3 confusion-matrix injection
        alone, applied to the validation split. This is equivalent to
        Phase~3 and serves as a reference point within the Phase~6 framework.

  \item \textbf{\texttt{self\_only}}: Phase~4 self-reflective prompting
        alone, applied to the validation split. This is equivalent to
        Phase~4 and serves as the second reference point.

  \item \textbf{\texttt{conf\_self}}: Phase~3 and Phase~4 combined.
        The system prompt includes both the failure-filtered confusion list
        (in \texttt{<confusion\_patterns>}) and the self-reflective diagnostic
        text with word-level exemplars (in \texttt{<self\_analysis>}).
        This tests whether the two knowledge types are complementary or
        redundant.

  \item \textbf{\texttt{best\_camel}}: The \texttt{conf\_self} prompt
        combination with Phase~5 CAMeL post-processing applied on top of the
        LLM output. This tests whether the morphological revert layer adds
        value on top of the best prompt-only combination.
\end{enumerate}

The combinations are chosen based on the isolated phase results: Phase~3 and
Phase~4 are the two strongest isolated strategies, and \texttt{conf\_self}
tests their combination directly. Phase~5 is then layered on top as an
orthogonal post-processing step, since it operates on the output rather than
the prompt.

\subsection{Ablation Design}
\label{subsec:phase6_ablation}

The ablation study measures the marginal contribution of each component by
removing it from the \texttt{conf\_self} combination and measuring the CER
impact. Two ablations are performed:

\begin{equation}\label{eq:ablation_conf}
  \Delta_{\text{CER}}^{(\text{conf})} = \text{CER}_{\text{self\_only}}
    - \text{CER}_{\text{conf\_self}}
\end{equation}

\begin{equation}\label{eq:ablation_self}
  \Delta_{\text{CER}}^{(\text{self})} = \text{CER}_{\text{conf\_only}}
    - \text{CER}_{\text{conf\_self}}
\end{equation}

A positive $\Delta_{\text{CER}}^{(k)}$ means removing component $k$ increases
CER --- the component was contributing positively. A negative value means
removing the component decreased CER --- the component was hurting when
combined.

\subsection{Statistical Testing}
\label{subsec:phase6_stats}

For Phase~6 results, paired t-tests are used to assess whether observed CER
differences between the best combination and Phase~2 are statistically
significant at the sample level. The Bonferroni correction is applied to
control the family-wise error rate across the four combination comparisons
at $\alpha = 0.05$. Bootstrap confidence intervals (1,000 iterations, seed 42)
are computed for the best-performing combination. Cohen's $d$ effect size is
reported alongside $p$-values to characterise the practical magnitude of
improvements. These tests are implemented in
\texttt{src/analysis/stats\_tester.py::StatsTester}.

\subsection{Output Files}
\begin{itemize}
  \item \texttt{results/phase6/\{combo\_id\}/\{dataset\}/metrics.json}
  \item \texttt{results/phase6/\{combo\_id\}/\{dataset\}/comparison\_vs\_phase2.json}
  \item \texttt{results/phase6/summary.json} --- all combinations ranked by
        PATS-A01 macro-average CER
  \item \texttt{results/phase6/ablation.json} --- ablation delta table
  \item \texttt{results/phase6/paper\_tables.md} --- formatted tables
\end{itemize}

\noindent
\textbf{Research question answered:} RQ6.

% ============================================================
\section{Phase Summary}
\label{sec:phase_summary}
% ============================================================

Table~\ref{tab:phase_summary} provides a consolidated reference for all phases.

\begin{table}[ht]
\caption{Summary of all experimental phases. ``Comparison'' is the phase each
         experiment is evaluated against. All phases are evaluated across nine
         datasets (eight PATS-A01 fonts + KHATT validation split).}
\label{tab:phase_summary}
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabularx}{\linewidth}{lXll}
\toprule
\textbf{Phase} & \textbf{Knowledge / Intervention} & \textbf{Comparison} & \textbf{RQ} \\
\midrule
Phase~1  & None (OCR analysis + taxonomy)         & N/A      & RQ1 \\
Phase~2  & None (zero-shot conservative prompt)   & Phase~1  & RQ2 \\
Phase~3  & Failure-filtered confusion matrix       & Phase~2  & RQ3 \\
Phase~4  & Self-reflective insights + word pairs  & Phase~2  & RQ4 \\
Phase~5  & CAMeL morphological revert + overcorrection revert & Phase~2 & RQ5 \\
Phase~6  & Combinations + ablation (4 configs)    & Phase~2 + best combo & RQ6 \\
\bottomrule
\end{tabularx}
\end{table}
