\setstretch{1.5}
\setlength{\parskip}{0.5\baselineskip}

This chapter describes the complete experimental infrastructure underlying the study.
The goal is to provide sufficient detail for full reproducibility: a reader should be
able to reconstruct the pipeline, datasets, preprocessing steps, and software modules
from this chapter alone.
Section~\ref{sec:arch} introduces the three-stage pipeline architecture.
Section~\ref{sec:datasets} describes the corpora used.
Section~\ref{sec:eval} covers preprocessing and evaluation.
Section~\ref{sec:kb} describes the knowledge base components used across phases.
Section~\ref{sec:llm_module} documents the LLM correction module.
Section~\ref{sec:promptbuilder} presents the \texttt{PromptBuilder} and all prompt
variants.

% ============================================================
\section{Overall Architecture}
\label{sec:arch}
% ============================================================

The primary constraint shaping system design is the absence of local GPU access.
The research workstation can load and process text data efficiently but cannot
run a 4-billion-parameter language model at useful inference speed.
At the same time, cloud GPU platforms such as Kaggle and Google Colab provide
free access to NVIDIA T4 GPUs that are well-suited for this model size.
The architecture is designed to exploit this resource split cleanly.

\subsection{Three-Stage Pipeline}
\label{subsec:pipeline}

The pipeline is divided into three stages that communicate through a single
intermediary file in JSON Lines format (JSONL).
Fig.~\ref{fig:pipeline} illustrates the flow.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  box/.style={rectangle, draw, rounded corners=4pt, minimum width=3.2cm,
              minimum height=1.1cm, align=center, font=\small},
  arrow/.style={-{Stealth[length=7pt]}, thick},
  lbl/.style={font=\scriptsize\itshape, text width=3cm, align=center}
]
  \node[box, fill=blue!10]   (export)  at (0,0)  {Export Stage\\{\footnotesize (local machine)}};
  \node[box, fill=green!10]  (infer)   at (5.5,0) {Inference Stage\\{\footnotesize (Kaggle / Colab GPU)}};
  \node[box, fill=orange!10] (analyze) at (11,0)  {Analyze Stage\\{\footnotesize (local machine)}};

  \draw[arrow] (export)  -- node[above,font=\scriptsize]{\texttt{inference\_input.jsonl}} (infer);
  \draw[arrow] (infer)   -- node[above,font=\scriptsize]{\texttt{corrections.jsonl}}      (analyze);

  \node[lbl, below=0.6cm of export]  {Load OCR + GT\\Build prompts\\Write JSONL};
  \node[lbl, below=0.6cm of infer]   {Load model\\Run prompts\\Append corrections};
  \node[lbl, below=0.6cm of analyze] {Read corrections\\Compute CER/WER\\Write reports};
\end{tikzpicture}
\caption{Three-stage pipeline architecture. Stages communicate via JSONL files on disk,
         enabling inference to run on any cloud GPU without requiring direct access to
         local data files.}
\label{fig:pipeline}
\end{figure}

\textbf{Stage 1 --- Export} runs on the local machine.
Each phase pipeline (e.g.\ \texttt{pipelines/run\_phase2.py}) loads the dataset,
pairs each OCR prediction with its ground-truth text, builds the phase-specific
prompt for each sample using \texttt{PromptBuilder}, and writes one JSON object
per sample to \texttt{inference\_input.jsonl}.
The file is stored under a phase- and dataset-specific results directory and is
transferred to the cloud environment by cloning the repository directly.

\textbf{Stage 2 --- Inference} runs on a Kaggle kernel or Google Colab notebook.
The script \texttt{scripts/infer.py} is the unified inference entry point for all
phases.
It loads the model and tokenizer once at startup, then reads
\texttt{inference\_input.jsonl} line by line, generates a correction for each
prompt, and appends the result to \texttt{corrections.jsonl}.
A resume mechanism reads the output file on startup and skips any sample whose
\texttt{sample\_id} already appears in it, making the inference step fully
restartable after an interruption without duplicating work.

\textbf{Stage 3 --- Analyze} runs on the local machine after inference completes.
The phase pipeline reads \texttt{corrections.jsonl}, computes CER and WER for each
sample by comparing the corrected text against the ground truth, aggregates
statistics across the dataset, and writes structured JSON reports and a
human-readable Markdown summary.

\subsection{JSONL Record Format}
\label{subsec:jsonl}

Each line in both JSONL files is a self-contained JSON object.
Table~\ref{tab:jsonl} lists all fields and their purpose.

\begin{table}[ht]
\caption{Fields in the inference JSONL files. All fields from the export stage are
         preserved through inference; the \texttt{correction} and status fields are
         appended by the inference script.}
\label{tab:jsonl}
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lll}
\toprule
\textbf{Field} & \textbf{Stage added} & \textbf{Description} \\
\midrule
\texttt{sample\_id}      & Export    & Unique identifier: \texttt{dataset:file\_stem} \\
\texttt{dataset}         & Export    & Dataset key, e.g.\ \texttt{PATS-A01-Akhbar-train} \\
\texttt{ocr\_text}       & Export    & Raw Qaari prediction string \\
\texttt{gt\_text}        & Export    & Ground-truth text (for analysis stage only) \\
\texttt{prompt}          & Export    & Serialised chat messages list (JSON array) \\
\texttt{prompt\_version} & Export    & Version tag, e.g.\ \texttt{p2v2}, \texttt{p3v1} \\
\texttt{correction}      & Inference & LLM output text \\
\texttt{success}         & Inference & \texttt{true} if generation succeeded; else fallback used \\
\texttt{latency\_s}      & Inference & Wall-clock generation time in seconds \\
\bottomrule
\end{tabular}
\end{table}

\noindent
This design has several practical advantages.
The inference machine needs no file-system access beyond the JSONL input file:
all ground-truth text, phase-specific context, and metadata are embedded in each
record at export time.
Adding a new phase requires only implementing a new \texttt{PromptBuilder} method
and a new pipeline script; the inference script and analysis logic are unchanged.
Finally, the JSONL format is human-readable and can be inspected with any text
editor or streamed through standard command-line tools for quick sanity checks.

\subsection{Resume and Force-Rerun Behaviour}
\label{subsec:resume}

All three stages implement resumable execution.
The export stage checks whether the target \texttt{inference\_input.jsonl} already
contains a record for a given dataset key before processing it, enabling partial
re-runs after adding new datasets.
The inference stage reads the output file on startup and populates a set of
completed \texttt{sample\_id} values; any sample already in the set is skipped.
The analyze stage checks whether the primary output file
(\texttt{results/<phase>/<dataset>/metrics.json}) already exists before
recomputing metrics.
All stages accept a \texttt{--force} flag to override resume behaviour and
reprocess everything from scratch.
This design prevents accidental data loss while keeping re-runs cheap when only
a subset of datasets changes.

% ============================================================
\section{Datasets}
\label{sec:datasets}
% ============================================================

Two Arabic text datasets are included in the study.
All quantitative experiments reported in this thesis use a single dataset:
PATS-A01, Akhbar font, training split, 200 samples.
The KHATT dataset is included in the data loading infrastructure but its
evaluation is deferred to future work.

\subsection{PATS-A01 --- Printed Arabic Text Set}
\label{subsec:pats}

PATS-A01 is a publicly available Arabic typewritten text recognition dataset.
It contains page images of Arabic text typeset in eight distinct fonts, together
with line-level and page-level ground truth.

\subsubsection{Fonts and Coverage}

The eight fonts included in PATS-A01 are:
Akhbar, Andalus, Arial, DecoType Naskh, Simplified Arabic, Tahoma, Thuluth,
and Traditional Arabic.
These fonts span a wide range of visual styles from the clean modern sans-serif
Tahoma to the decorative classical Thuluth, providing diversity in glyph shape
and rendering complexity that elicits different OCR error patterns across fonts.
All experiments in this thesis use the Akhbar font, which is a common newspaper
display font representative of modern typewritten Arabic.

\subsubsection{Ground-Truth Encoding}

Ground-truth files are encoded in Windows Arabic Code Page 1256 (CP-1256), a
legacy single-byte encoding used widely in Arabic software on Windows systems.
The DataLoader (\texttt{src/data/data\_loader.py}) converts all ground-truth text
to UTF-8 at load time before any downstream processing.
The Traditional font uses a differently named ground-truth file
(\texttt{TraditionalText.txt}) instead of the naming convention followed by the
other seven fonts (\texttt{A01-\{font\}Text.txt}); the DataLoader handles this
as a special case.

\subsubsection{Train/Validation Split}

PATS-A01 does not provide an official train/validation split.
An 80/20 random split was generated using the script
\texttt{scripts/generate\_pats\_splits.py} with a fixed random seed (42) for
reproducibility.
The resulting split is stored as a JSON file at
\texttt{data/ocr-raw-data/PATS\_A01\_Dataset/pats\_splits.json}
and records sample identifiers assigned to either the training or validation
partition.
Each font yields 2,213 training samples and 553 validation samples
(2,766 total per font).
Each dataset configuration entry uses the field
\texttt{pats\_split: "train"} or \texttt{pats\_split: "validation"}
to select the appropriate partition at load time, ensuring that experiments
that evaluate on training data are never accidentally contaminated with
validation samples.

\subsubsection{Scope of This Thesis}

All quantitative experiments use the Akhbar font training split.
A random sample of 200 items is drawn from this split for all phase comparisons,
providing a practical compute budget (each of the 200 samples requires one model
inference call) while remaining large enough for meaningful statistical analysis.
Phase~4D additionally uses the validation split to analyse error patterns on
held-out data, as described in Chapter~4.

\subsection{KHATT --- Khatt Arabic Handwritten Text}
\label{subsec:khatt}

KHATT is an Arabic handwritten text dataset \cite{mahmoud2014khatt}.
It contains text written by multiple writers in unconstrained natural handwriting,
representing realistic variation in stroke style, baseline angle, character
connectivity, and ink characteristics.
The dataset provides separate training and validation splits.
Within this project, Qaari results for KHATT are stored under
\texttt{data/ocr-results/qaari-results/khatt-data/\{split\}/\{Training|Validation\}/}
and the DataLoader supports them using the keys \texttt{KHATT-train} and
\texttt{KHATT-validation}.
Quantitative evaluation on KHATT is deferred to future work because the primary
research question --- whether prompt augmentation improves correction on
high-quality typewritten Arabic --- is best studied on a controlled, homogeneous
dataset first.
KHATT is expected to present a more challenging OCR baseline due to unconstrained
handwriting, and it is plausible that knowledge augmentation strategies would
behave differently in that setting.

\subsection{Dataset Statistics}
\label{subsec:dataset_stats}

Table~\ref{tab:datasets} summarises all datasets included in the system.

\begin{table}[ht]
\caption{Dataset summary. ``Evaluated'' is the number of samples used in experiments
         reported in this thesis. ``---'' indicates that statistics depend on
         writer count and session structure not yet fully catalogued.}
\label{tab:datasets}
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabularx}{\linewidth}{lllrrl}
\toprule
\textbf{Dataset Key} & \textbf{Type} & \textbf{Split} & \textbf{Total} & \textbf{Evaluated} & \textbf{Notes} \\
\midrule
PATS-A01-Akhbar-train      & Typewritten & Train      & 2,213 & 200 & Primary evaluation \\
PATS-A01-Akhbar-val        & Typewritten & Validation & 553   & 200 & Phase~4D only \\
PATS-A01-Andalus-\{train|val\} & Typewritten & Both   & 2,766 & 0   & Future work \\
PATS-A01-Arial-\{train|val\}   & Typewritten & Both   & 2,766 & 0   & Future work \\
PATS-A01-Naskh-\{train|val\}   & Typewritten & Both   & 2,766 & 0   & Future work \\
PATS-A01-Simplified-\{train|val\} & Typewritten & Both & 2,766 & 0  & Future work \\
PATS-A01-Tahoma-\{train|val\}  & Typewritten & Both   & 2,766 & 0   & Future work \\
PATS-A01-Thuluth-\{train|val\} & Typewritten & Both   & 2,766 & 0   & Future work \\
PATS-A01-Traditional-\{train|val\} & Typewritten & Both & 2,766 & 0 & Future work \\
KHATT-train                & Handwritten & Train      & ---   & 0   & Future work \\
KHATT-validation           & Handwritten & Validation & ---   & 0   & Future work \\
\midrule
\multicolumn{4}{l}{\footnotesize Total PATS-A01 font-split combinations: 16 (8 fonts $\times$ 2 splits)} & & \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{OCR Results Format}
\label{subsec:ocr_format}

Qaari produces one plain-text \texttt{.txt} file per input image.
Each file contains the recognised text for that document page.
OCR results are stored under \texttt{data/ocr-results/qaari-results/} with
sub-directories organised by dataset type:
\begin{itemize}
  \item PATS-A01: \texttt{qaari-results/pats-a01-data/A01-\{font\}/}
  \item KHATT:    \texttt{qaari-results/khatt-data/\{split\}/\{Training|Validation\}/}
\end{itemize}

\noindent
The DataLoader pairs each OCR file with its corresponding ground-truth entry by
matching the file stem.
The OCR engine is configurable: the configuration field \texttt{data.ocr\_model}
names the sub-directory within \texttt{data/ocr-results/} that holds all results
for that engine.
Changing this field switches the entire pipeline to a different OCR engine's output
directory without any code modification.
This design anticipates future comparative studies across multiple Arabic OCR engines.

% ============================================================
\section{Preprocessing and Evaluation}
\label{sec:eval}
% ============================================================

\subsection{Text Normalisation}
\label{subsec:norm}

Text normalisation is applied identically to both the hypothesis (OCR output or
LLM correction) and the reference (ground truth) before any metric is computed.
The following transformations are applied:

\begin{enumerate}
  \item \textbf{Diacritic stripping.}
        Arabic short vowel marks (harakat/tashkeel) occupy Unicode code points
        U+064B through U+065F and related ranges (U+0610--U+061A for extended
        marks, U+0670 for the superscript alef).
        These characters are removed from both texts before computing CER and WER.
        The rationale is that diacritics are stylistically optional in most
        modern Arabic text; their presence or absence does not change lexical
        identity.
        Penalising a corrector for changing a diacritised form to the same word
        without diacritics, or for failing to add diacritics that were absent in
        the OCR input, would conflate orthographic convention with correction
        quality.
        Both the raw metric (with diacritics) and the normalised metric (without
        diacritics) are computed and stored in all result files;
        the diacritic-stripped variant is reported as primary throughout this thesis.

  \item \textbf{Tatweel removal.}
        The Arabic elongation character tatweel (kashida, U+0640) is a decorative
        stroke that extends word shapes without changing lexical identity.
        It is removed from both texts before metric computation.

  \item \textbf{No letter normalisation.}
        Alef variants, hamza forms, and the alef maqsura / ya distinction are
        \emph{not} normalised before evaluation.
        Collapsing these distinctions (e.g.\ mapping all alef variants to a
        bare alef) would mask a class of errors that the study specifically
        aims to measure.
        The confusion matrix and error taxonomy both depend on the uncollapsed
        character set.
\end{enumerate}

\noindent
Normalisation is implemented in \texttt{src/data/text\_utils.py::normalise\_arabic()}.

\subsection{CER and WER Computation}
\label{subsec:metrics}

\subsubsection{Definitions}

Character Error Rate (CER) measures the minimum edit distance between the hypothesis
string $h$ and the reference string $r$ at the character level, normalised by the
length of the reference:
\begin{equation}\label{eq:cer}
  \mathrm{CER} = \frac{S_c + D_c + I_c}{N_c}
\end{equation}
where $S_c$, $D_c$, and $I_c$ are the number of character-level substitutions,
deletions, and insertions in the minimum-edit alignment, and $N_c$ is the total
number of characters in the reference.
Word Error Rate (WER) applies the same formula at the word level, treating
whitespace-delimited tokens as the atomic unit:
\begin{equation}\label{eq:wer}
  \mathrm{WER} = \frac{S_w + D_w + I_w}{N_w}
\end{equation}
CER is the primary metric throughout this thesis.
Arabic morphology makes word-level metrics more sensitive to tokenisation choices;
character-level errors more directly capture the visual confusion phenomena (dot
placement, shape similarity) that are the dominant error type in Arabic OCR.

\subsubsection{Implementation}

CER is computed using the \texttt{editdistance} library, which provides an
efficient implementation of the Levenshtein distance.
WER is computed using the \texttt{jiwer} library.
Both are computed \emph{per sample first}, then aggregated:
\begin{equation}\label{eq:sample_cer}
  \mathrm{CER}_i = \frac{\mathrm{editdistance}(h_i,\, r_i)}{\max(|r_i|, 1)}
\end{equation}
where $|r_i|$ is the reference character count for sample $i$.
Using per-sample CER (rather than pooling all characters across the corpus) ensures
that every sample contributes equally to the aggregate mean regardless of its length.
This prevents the few long documents in the dataset from dominating the statistic.

Aggregate statistics reported for each dataset-phase combination are:
mean CER ($\mu$), standard deviation ($\sigma$), median, and 95th percentile (p95).
The standard deviation quantifies variability across samples; the p95 captures
worst-case performance and is particularly useful for identifying the impact of
catastrophic OCR failures such as the repetition artifact described in Chapter~5.

\subsection{Error Taxonomy Construction}
\label{subsec:taxonomy}

Phase~1 classifies each substitution error into one of seven categories based on
the character pair (ground-truth character, OCR character) found in the minimum-edit
alignment between ground-truth and OCR text.
Table~\ref{tab:taxonomy} defines each category and its detection criterion.
Categories are applied in order; a substitution that matches an earlier category
is not re-counted in a later one.

\begin{table}[ht]
\caption{Error taxonomy used in Phase~1 baseline analysis. Categories are ordered
         from most specific (dot confusion) to most general (other substitution).}
\label{tab:taxonomy}
\centering
\renewcommand{\arraystretch}{1.35}
\begin{tabularx}{\linewidth}{lX}
\toprule
\textbf{Category} & \textbf{Detection criterion} \\
\midrule
Dot confusion
  & Substitution between characters that share the same base glyph and differ
    only in the number or position of diacritical dots:
    \{ba, ta, tha, nun, ya\},
    \{qaf, fa\},
    \{jim, ha, kha\},
    \{dal, dhal\}. \\
Hamza errors
  & Substitution among alef variants or between hamza forms:
    alef, alef with hamza above, alef with hamza below, alef madda,
    hamza alone, hamza on waw seat, hamza on ya seat. \\
Taa marbuta
  & Substitution between taa marbuta (feminine ending) and ha,
    which share the same base shape and differ only by two dots. \\
Alef maqsura
  & Substitution between alef maqsura (word-final /a/ written with a
    ya-shaped letter without dots) and ya (with dots). \\
Similar shapes
  & Substitution between characters sharing the same skeleton but differing
    in secondary visual features:
    \{ra, zay\}, \{dal, dhal\}, \{ta marbuta, ha\},
    \{ta with extended tail (\RL{Ř}), dha (\RL{Ů})\}. \\
Merged words
  & Fewer words in the OCR output than in the ground truth;
    a word boundary is missing. \\
Split words
  & More words in the OCR output than in the ground truth;
    a spurious word boundary is inserted. \\
Insertion / Deletion / Other
  & All remaining edit operations: character insertions, character deletions,
    and substitutions not classified by any of the above categories. \\
\bottomrule
\end{tabularx}
\end{table}

% ============================================================
\section{Knowledge Base Components}
\label{sec:kb}
% ============================================================

Phases~3 through~5 (and Phase~6) each inject a different type of external knowledge
into the LLM prompt.
All knowledge loaders are implemented in \texttt{src/data/knowledge\_base.py}
and share a common interface: a \texttt{load()} method that reads source data
and a \texttt{format\_for\_prompt()} method that returns a plain Arabic string
ready for insertion into a system prompt.

\subsection{Confusion Matrix Loader}
\label{subsec:kb_confusion}

\texttt{ConfusionMatrixLoader} reads the \texttt{confusion\_matrix.json} file
produced by Phase~1.
The confusion matrix records, for every ground-truth character, the frequency
with which Qaari substituted each other character in its place.

The loader supports two code paths.
If the Phase~1 output contains a pre-computed \texttt{top\_20} sorted list, it is
used directly for efficiency.
Otherwise, the full confusion dictionary is flattened and sorted by substitution
count.
The \texttt{format\_for\_prompt()} method selects the top-$N$ confusion pairs
(default $N = 10$, configured via \texttt{phase3.top\_n} in
\texttt{configs/config.yaml}) and formats them as an Arabic-language bullet list,
where each item identifies a source character and its most common Qaari substitute.

A dataset with fewer than 200 total substitutions (threshold configurable as
\texttt{phase3.min\_substitutions}) falls back to a pooled confusion matrix
aggregated across all datasets of the same type (e.g.\ all PATS-A01 training
datasets).
This prevents noise-dominated confusion matrices from individual small datasets
from distorting the injection content.

\subsection{Rules Loader}
\label{subsec:kb_rules}

\texttt{RulesLoader} loads Arabic orthographic rules from the directory
\texttt{data/rules/}, which contains JSONL files (one rule object per line) with
fields for the rule text, its category, and example words illustrating the rule.
The loader selects rules across three files: orthography, morphology, and syntax.
The \texttt{format\_for\_prompt()} method formats the selected rules as a
numbered instruction list in Arabic.

Table~\ref{tab:rules} lists the primary orthographic rule categories injected in
Phase~4A.
These categories were chosen because they correspond to the error types most
likely to appear in Arabic typewritten OCR output, as confirmed by the Phase~1
confusion matrix and error taxonomy.

\begin{table}[ht]
\caption{Arabic orthographic rule categories used in Phase~4A. Each category
         targets a specific phenomenon that both human writers and OCR engines
         commonly mishandle.}
\label{tab:rules}
\centering
\renewcommand{\arraystretch}{1.35}
\begin{tabularx}{\linewidth}{lX}
\toprule
\textbf{Rule category} & \textbf{Description and OCR relevance} \\
\midrule
Hamza al-qat' (cutting hamza)
  & Hamza written visibly above or below the alef in all positions.
    OCR frequently drops the hamza diacritic and outputs a bare alef. \\
Hamza al-wasl (conjunctive hamza)
  & An alef at the start of certain verb forms and nouns whose hamza is
    elided in connected speech and not written.
    Distinguishing these two types governs the spelling of the definite
    article (\textit{al-}), some verbal nouns, and Arabic proper nouns. \\
Taa marbuta vs.\ ha
  & The feminine ending taa marbuta and the letter ha share the same base
    shape; OCR dot-placement errors cause frequent confusion between them. \\
Alef maqsura vs.\ ya
  & Word-final /a/ written as alef maqsura (a ya-shaped letter without dots)
    is frequently confused with ya (with two dots below) by OCR. \\
Tanwin (nunation)
  & Double-diacritic marking for indefinite nouns; the accusative tanwin
    al-fath is written as a doubled fatha above an alef, which OCR may
    misread or drop. \\
Definite article assimilation
  & Sun letters assimilate the lam of the definite article (\textit{al-});
    moon letters do not.
    This governs whether the written lam appears explicitly in the word. \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{QALB Loader and Filtering}
\label{subsec:kb_qalb}

\texttt{QALBLoader} reads the Qatar Arabic Language Bank (QALB) corpus from
\texttt{data/QALB-0.9.1-Dec03-2021-SharedTasks/}.
QALB contains human-authored Arabic sentences paired with manually corrected
versions, annotating grammatical, spelling, and punctuation errors.
The 2014 training split is used in this study.

After loading, the following filters are applied in sequence:

\begin{enumerate}
  \item \textbf{Length filter.}
        Pairs with source sentences shorter than 10 characters or longer than
        300 characters are discarded.
        The upper bound of 300 characters was determined empirically:
        QALB sentences average between 161 and 594 characters, so a stricter
        original default of 100 characters eliminated the entire corpus.

  \item \textbf{Word-change filter.}
        Pairs where the correction modifies more than 15 words are discarded.
        Heavily-edited sentences are unsuitable as demonstrations because they
        may signal to the model that aggressive rewriting is expected.

  \item \textbf{OCR-relevance filter.}
        A pair is retained only if its error type overlaps with known OCR error
        patterns: dot-group substitutions, alef/hamza variants, taa marbuta / ha
        swaps, or similar-shape confusions.
        Pairs that correct purely grammatical or agreement errors without any
        spelling-level change are discarded, as they could teach the model to
        make rewrites that OCR would never introduce.
\end{enumerate}

\noindent
After all three filters, 528 OCR-relevant pairs are available.
At inference time, Phase~4B selects 5 pairs using a diversity-maximising strategy
(\texttt{selection: "diverse"}) that ensures the selected examples span multiple
error categories rather than repeating the same phenomenon.

\subsection{OpenITI Corpus and RAG Index}
\label{subsec:kb_openiti}

\texttt{OpenITILoader} extracts clean Arabic sentences from the Open Islamicate
Texts Initiative (OpenITI) corpus at \texttt{data/OpenITI/}.
OpenITI is a large-scale digitised collection of Islamicate texts spanning
classical Arabic, Persian, and other languages of the Islamic scholarly tradition,
containing hundreds of millions of words of Arabic text.
It was chosen for RAG because of its size, which allows construction of a large
and varied retrieval index.

\subsubsection{Corpus Processing}

The loader reads an inventory CSV file that records the local path and metadata
of each OpenITI text file.
Each text file is split into candidate sentences at punctuation boundaries
(full stop, question mark, exclamation mark, and the Arabic semicolon U+061B).
Sentences shorter than 30 characters or longer than 300 characters are discarded.
Up to 200,000 sentences are sampled from primary-status texts (those with
\texttt{status: "pri"} in OpenITI metadata), using a fixed random seed (42).

\subsubsection{FAISS Retrieval Index}

The filtered sentences are embedded using the multilingual sentence encoder
\texttt{sentence-transformers/paraphrase-multilingual-MiniLM-L12-v2},
an efficient 12-layer model covering over 50 languages including Arabic.
Embeddings are computed in batches of 256 and indexed in a FAISS flat
inner-product index (\texttt{FlatIP}).
The index is built once locally by running
\texttt{python pipelines/run\_phase5.py --mode build} and saved to disk.

At inference time, each OCR input text is embedded as a query and the top-3
most similar sentences are retrieved by inner-product similarity.
These sentences are formatted as a numbered Arabic list and injected into the
system prompt as grounding context.
An important caveat on domain match is discussed in Chapter~5: OpenITI contains
classical and historical Arabic, while the PATS-A01 Akhbar dataset contains modern
newspaper text.
The semantic overlap between these domains is low, which affects the quality of
retrieved passages.

\subsection{CAMeL Tools Integration}
\label{subsec:kb_camel}

CAMeL Tools is an open-source Arabic NLP toolkit developed at the Computational
Approaches to Modeling Language (CAMeL) Lab, NYU Abu Dhabi \cite{obeid2020camel}.
This study uses its morphological analysis component, backed by the Modern Standard
Arabic database \texttt{calima-msa-r13}.

\subsubsection{MorphAnalyzer}

\texttt{src/linguistic/morphology.py::MorphAnalyzer} wraps the CAMeL Tools
\texttt{MorphologicalAnalyzer} with two extensions:

\begin{itemize}
  \item \textbf{LRU cache.}
        The morphological analyser is computationally expensive per lookup.
        Repeated queries for the same token are served from a Least-Recently-Used
        cache of size 10,000, avoiding redundant analyser calls for frequently
        occurring words.

  \item \textbf{Graceful degradation.}
        If \texttt{camel\_tools} is not installed or the morphology database has
        not been downloaded, \texttt{MorphAnalyzer.analyse()} returns \texttt{None}
        rather than raising an exception.
        Downstream code that checks for \texttt{None} can continue in a degraded
        mode without CAMeL-based functionality.
\end{itemize}

\subsubsection{WordValidator and the Revert Strategy}

\texttt{src/linguistic/validator.py::WordValidator} implements the Phase~4C
post-processing logic.
For each word token in the LLM-corrected output, it applies the following
decision rule:

\begin{quote}
If the LLM correction for a word \emph{fails} morphological analysis (zero valid
analyses) \emph{and} the corresponding OCR word \emph{passes} morphological
analysis (at least one valid analysis), revert to the OCR word.
\end{quote}

\noindent
The strategy is deliberately conservative.
It can only revert a word to the OCR original; it cannot introduce a new word
that neither the OCR nor the LLM produced.
Consequently, the revert strategy has no downside risk on any word it touches,
but its upside is limited to cases where the LLM introduces a morphologically
invalid Arabic form while the OCR had a morphologically valid one.
The validator records summary statistics for each text: total Arabic tokens
compared, reverted count, and revert rate.

\subsection{LLM Insights Loader}
\label{subsec:kb_insights}

\texttt{LLMInsightsLoader} reads JSON files from
\texttt{results/phase4d/insights/}, which are produced during Phase~4D's
training-split analysis step.
Each file records, for each error category, two statistics derived from comparing
Phase~2 LLM corrections against ground truth on the training split:
\begin{itemize}
  \item \texttt{fix\_rate}: the fraction of OCR errors of that category that
        the LLM correctly fixed (higher is better).
  \item \texttt{introduction\_rate}: the fraction of ground-truth tokens of that
        category that the LLM \emph{incorrectly} changed (lower is better).
\end{itemize}
The loader converts these statistics into Arabic-language diagnostic sentences
using the following logic:
error types with \texttt{fix\_rate} $> 0.6$ are reported as strengths;
those with \texttt{fix\_rate} $< 0.4$ are reported as weaknesses;
those with \texttt{introduction\_rate} $> 0.05$ are flagged as over-correction
risks.
Statistics are aggregated at the dataset-type level (PATS-A01 vs.\ KHATT) to
smooth per-sample noise.
Up to three weaknesses and two over-correction warnings are injected into the
Phase~4D system prompt (configurable thresholds in \texttt{phase4d.insights}).

% ============================================================
\section{The LLM Correction Module}
\label{sec:llm_module}
% ============================================================

\subsection{Class Hierarchy}
\label{subsec:corrector_arch}

All LLM backend implementations share a common abstract interface defined in
\texttt{src/core/llm\_corrector.py}.
Fig.~\ref{fig:corrector} shows the class hierarchy.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  cls/.style={rectangle, draw, rounded corners=3pt, minimum width=4.5cm,
              minimum height=0.9cm, align=center, font=\small},
  inh/.style={-{Triangle[open]}, thick}
]
  \node[cls, fill=gray!15] (base) at (0,0) {\texttt{BaseLLMCorrector} (abstract)};
  \node[cls, fill=blue!10] (tc)  at (-4,-2.5) {\texttt{TransformersCorrector}};
  \node[cls, fill=green!10](mock) at (0,-2.5) {\texttt{MockCorrector}};
  \node[cls, fill=orange!10](api) at (4,-2.5)  {\texttt{APICorrector}};

  \draw[inh] (tc.north)   -- ++(0,0.7) -| (base.south);
  \draw[inh] (mock.north) -- (base.south);
  \draw[inh] (api.north)  -- ++(0,0.7) -| (base.south);
\end{tikzpicture}
\caption{LLM corrector class hierarchy. All pipeline code depends only on the
         \texttt{BaseLLMCorrector} interface; the concrete backend is selected
         at runtime by \texttt{get\_corrector(config)}.}
\label{fig:corrector}
\end{figure}

\textbf{\texttt{BaseLLMCorrector}} is an abstract base class that defines two
abstract methods: \texttt{correct()} and \texttt{model\_name}.
Implementations must satisfy the contract that \texttt{correct()} never raises
an exception and always returns a \texttt{CorrectionResult} with a non-empty
\texttt{corrected\_text} field.
On any failure, \texttt{corrected\_text} is set to the original
\texttt{ocr\_text} (a safe fallback that does not inflate error rates beyond
the OCR baseline).

\textbf{\texttt{TransformersCorrector}} is the production backend.
It loads the model and tokenizer via HuggingFace Transformers once at startup
and reuses them for all subsequent inference calls.
It is designed to run on Kaggle or Colab GPU kernels.
Optional 4-bit quantization via \texttt{bitsandbytes} is available for GPUs
with less than 8~GB of VRAM, but was not used in the experiments reported here
(the Kaggle T4 GPU has 16~GB).

\textbf{\texttt{MockCorrector}} returns the OCR text unchanged.
It requires no model, no GPU, and no internet access.
It is activated by setting \texttt{model.backend: "mock"} and is used for
local pipeline validation: a full export--infer--analyze cycle completes
instantaneously with mock, producing valid JSON outputs with CER equal to the
OCR baseline (since no correction is performed).

\textbf{\texttt{APICorrector}} provides an OpenAI-compatible REST interface for
potential future use with hosted models.
It is not used in the experiments of this thesis.

\subsection{Generation Configuration}
\label{subsec:gen_config}

Table~\ref{tab:gen_config} documents the exact generation parameters used in all
phases of this study.
All phases use identical generation settings; only the prompt content changes
between phases.

\begin{table}[ht]
\caption{Generation parameters for \texttt{TransformersCorrector}.}
\label{tab:gen_config}
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lll}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Rationale} \\
\midrule
Model ID             & \texttt{Qwen/Qwen3-4B-Instruct-2507} & Primary study model \\
\texttt{max\_new\_tokens} & 1024        & Sufficient headroom for all prompt types \\
\texttt{do\_sample}       & \texttt{True}  & Stochastic sampling \\
\texttt{temperature}      & 0.1         & Near-deterministic; suppresses high-entropy output \\
\texttt{enable\_thinking} & \texttt{False} & Suppresses Qwen3 chain-of-thought scratchpad \\
\texttt{MAX\_INPUT\_TOKENS} & 4,096    & Out-of-memory guard on unexpectedly long prompts \\
\bottomrule
\end{tabular}
\end{table}

\noindent
Temperature 0.1 is chosen to make outputs nearly deterministic across runs while
avoiding the degenerate behaviour that pure greedy decoding can produce.
At this temperature the output distribution is heavily peaked at the
most-probable token at each step, but a small amount of stochasticity is retained
to prevent the model from entering repetitive loops when correcting ambiguous text.

The \texttt{enable\_thinking} flag is Qwen3-specific.
When enabled, Qwen3 prepends a chain-of-thought scratchpad enclosed in
\texttt{<think>...</think>} tokens before generating the final answer.
For Arabic post-OCR correction, a one-pass text-to-text transformation task,
the scratchpad adds latency and output tokens without improving the correction
output.
More critically, the scratchpad contents are sometimes included in the decoded
output, breaking the response extraction step that expects only the corrected
text.
Setting \texttt{enable\_thinking=False} in \texttt{apply\_chat\_template()}
prevents this.

\subsection{Response Extraction}
\label{subsec:extraction}

The model is instructed to return \emph{only} the corrected text, with no
explanation or commentary.
Despite this instruction, decoded outputs occasionally contain preamble or
non-Arabic content.
The method \texttt{\_extract\_corrected\_text()} applies the following steps:

\begin{enumerate}
  \item Strip leading and trailing whitespace and newlines.
  \item If the result is empty, log a warning and return the original OCR text.
  \item If the result contains no Arabic characters (Unicode block U+0600--U+06FF),
        the model has likely replied in English or refused the request; log a
        warning and return the original OCR text.
  \item Otherwise, return the cleaned string as the correction.
\end{enumerate}

\noindent
Steps~2 and~3 are safety checks that prevent model refusals or empty generations
from propagating as corrections.
Without them, a refusal would produce a CER of approximately 1.0 for that sample,
artificially degrading aggregate performance.
By falling back to the OCR text on failure, the method ensures that failed samples
do not inflate aggregate CER above the OCR baseline.
If \texttt{max\_retries} additional attempts all fail, the fallback is applied and
the \texttt{success} flag in the JSONL record is set to \texttt{false}.

% ============================================================
\section{PromptBuilder}
\label{sec:promptbuilder}
% ============================================================

\texttt{src/core/prompt\_builder.py::PromptBuilder} is the single point of contact
for all prompt construction in the study.
Each experimental phase has exactly one corresponding \texttt{build\_*} method.
All methods return a list of dictionaries in OpenAI chat format --- a system
message followed by a user message --- which is directly accepted by the HuggingFace
\texttt{apply\_chat\_template()} API.

\subsection{Prompt Design Philosophy}
\label{subsec:prompt_design}

All prompts in this study are written entirely in Arabic.
The decision to use Arabic rather than English for system messages was motivated
by two considerations.
First, the correction task is inherently Arabic-specific; using the target language
avoids any cross-lingual ambiguity in instruction following.
Second, Qwen3, as a multilingual model, was trained on Arabic instructions and
responds more reliably when the instruction language matches the text language.

All phase prompts share a \emph{conservative instruction}: the system message
tells the model that the input text may already be correct, that it should return
the text unchanged if it finds no clear errors, and that it must not change what
is already correct.
This instruction counteracts a known failure mode of instruction-following LLMs:
they tend to interpret any correction directive as a mandate to rewrite, even when
the input is already accurate.
The conservative constraint is particularly important in the PATS-A01 Akhbar
setting, where the OCR baseline CER on normal samples is only 1.37\% and the
vast majority of tokens require no change.
The conservative v2 zero-shot prompt was adopted after pilot runs showed that the
original aggressive v1 prompt (``correct this text'') led to unacceptably high
error rates on clean samples.

\subsection{Prompt Variants}
\label{subsec:prompt_variants}

Table~\ref{tab:prompts} provides a complete registry of all prompt variants.

\begin{table}[ht]
\caption{Prompt variant registry. Each phase uses exactly one variant.
         All prompts are in Arabic. Full prompt text is given in Appendix~A.}
\label{tab:prompts}
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabularx}{\linewidth}{llXl}
\toprule
\textbf{Method} & \textbf{Tag} & \textbf{Injected context} & \textbf{Phase} \\
\midrule
\texttt{build\_zero\_shot()}       & p2v2  & None (instruction only)              & 2  \\
\texttt{build\_ocr\_aware()}       & p3v1  & Top-10 Qaari character confusion pairs & 3  \\
\texttt{build\_rule\_augmented()}  & p4av1 & Arabic orthographic rules            & 4A \\
\texttt{build\_few\_shot()}        & p4bv1 & 5 QALB error-correction examples     & 4B \\
\texttt{build\_self\_reflective()} & p4dv1 & LLM error statistics from training split & 4D \\
\texttt{build\_rag()}              & p5v1  & Top-3 retrieved OpenITI sentences    & 5  \\
\texttt{build\_combined()}         & p6v1  & Any subset of the above contexts     & 6  \\
\bottomrule
\end{tabularx}
\end{table}

\subsubsection{Zero-Shot Prompt (p2v2)}

The zero-shot prompt consists of an Arabic system message and the raw OCR text
as the user message.
The system message establishes the model as an Arabic text proofreader and
explicitly states the conservative instruction: the OCR output may be fully
correct; the model should return it exactly if no clear error is found; it must
not change what is already correct; it should return only the text with no
explanation.
There is no injected external knowledge; the model's only guidance comes from
its own training and from the instruction framing.

\subsubsection{OCR-Aware Prompt (p3v1)}

The OCR-aware prompt extends the zero-shot system message with an injected
Arabic bullet list of confusion pairs from \texttt{ConfusionMatrixLoader}.
Each bullet item names a ground-truth character and the character Qaari commonly
substitutes for it.
The system message instructs the model to pay particular attention to these
character pairs when correcting.
If no confusion context is available, \texttt{build\_ocr\_aware()} falls back
transparently to the zero-shot prompt.

\subsubsection{Rule-Augmented Prompt (p4av1)}

The rule-augmented prompt replaces the confusion list with a numbered list of
Arabic orthographic rules from \texttt{RulesLoader}.
The system message instructs the model to observe these rules during correction.
The rules address the same phenomena targeted by the confusion matrix ---
dot-group characters, hamza, taa marbuta, alef maqsura --- but frame them as
prescriptive grammar rules rather than as statistical error patterns.

\subsubsection{Few-Shot Prompt (p4bv1)}

The few-shot prompt inserts five input--output demonstration pairs from the QALB
corpus into the system message.
Each pair is formatted as:
``Input: [erroneous sentence] / Correction: [corrected sentence]'' in Arabic.
The system message instructs the model to correct the new text following the same
approach as the examples.
The examples are selected by \texttt{QALBLoader} using the diversity strategy
to span multiple error types.

\subsubsection{Self-Reflective Prompt (p4dv1)}

The self-reflective prompt injects Arabic diagnostic sentences derived from the
model's own training-split error statistics.
The system message frames the diagnostics as performance feedback:
the model is informed which error categories it handles well, which it tends to
miss, and for which categories it tends to over-correct previously correct tokens.
This is a lightweight, pre-computed form of meta-cognitive guidance that avoids
the latency cost of asking the model to self-critique at inference time.

\subsubsection{RAG Prompt (p5v1)}

The RAG prompt injects three sentences retrieved from the OpenITI FAISS index.
The retrieved sentences are presented as reference Arabic text that the model
should use as grounding context.
The system message states that these reference sentences are correct Arabic text
similar to the input and that the model should consult them when deciding how to
correct the input.

\subsubsection{Combined Prompt (p6v1)}

\texttt{build\_combined()} accepts any subset of the five context types and
assembles them into a single system message using a fixed section ordering
numbered with Arabic ordinals (``First:\ confusions, Second:\ rules,
Third:\ examples, Fourth:\ retrieved passages, Fifth:\ performance notes'').
Only non-empty context strings are included; the closing conservative instruction
is always appended.
If all context strings are empty, the method falls back to the zero-shot prompt.
This design makes Phase~6 combination experiments straightforward to specify and
fully reproducible from the combination identifier alone.
