\setstretch{1.5}

%----------------------------------------------------------------------------
\section{The Problem with Arabic OCR}
\label{sec:intro_problem}
%----------------------------------------------------------------------------

The digitisation of Arabic text is a prerequisite for a broad class of
downstream applications: full-text search over historical archives, neural
machine translation of Arabic news and literature, text-to-speech systems
for visually impaired users, and information extraction from scanned
government documents. The technology that converts images of text into
machine-readable character sequences --- optical character recognition (OCR)
--- is the entry point for all of these applications. When OCR fails, the
failure propagates: a misread character changes word meaning, a sequence of
recognition errors degrades translation quality, and a corrupted document
image cannot be retrieved by keyword search.

Arabic presents challenges that English OCR does not face. The Arabic script
is cursive: letters connect within words and their shapes change depending on
their position in the word (isolated, initial, medial, or final). There are
no clear boundaries between characters within a word, so segmentation and
recognition are tightly coupled --- a segmentation error directly produces a
recognition error. Many Arabic letters share the same base skeleton and
differ only in the number and placement of small dots: the pair \RL{ب}/\RL{ي}
(ba/ya) differs only in whether the dot is below or above the baseline;
\RL{ف}/\RL{ق} (fa/qaf) differs only in having one or two dots above. Ink
irregularities, low scan resolution, or font variation can shift a dot from
its correct position and cause the OCR engine to output the wrong letter.

Short vowels (diacritics, or \textit{tashkeel}) are generally optional in
modern printed Arabic and are absent from most typewritten text, yet their
presence or absence changes word meaning and complicates matching between
OCR output and ground truth. Classical and modern standard Arabic coexist
with several regional dialects, each with partially distinct orthographies.
Ligatures --- special glyphs for common letter combinations --- further
increase the visual complexity that OCR engines must resolve.

A further complication is the diversity of fonts used in printed Arabic.
Each typeface imposes its own glyph shapes, dot placement conventions,
stroke widths, and ligature styles. A font with decorative rendering may
visually merge characters that a plainer font keeps distinct; a calligraphic
font may exaggerate strokes to the point where letter identity becomes
ambiguous even to human readers at a glance. This font-induced variation
means that the same underlying text produces substantially different
character-level error patterns depending on the typeface used to render it.

The consequence of these challenges is that open-source Arabic OCR engines
produce recognisable error patterns. Character-level confusions between
visually similar letter pairs, word-level segmentation failures, and
occasional catastrophic output loops inflate the character error rate (CER)
in ways that downstream applications cannot silently absorb. A single
confused character in a word can change its meaning entirely; a repeated
output loop corrupts an entire document page. These are not random errors;
they are systematic, rooted in the visual properties of the Arabic script
and in the specific training data and architecture of each OCR engine. Their
systematic nature motivates the hypothesis that a language model, given the
erroneous OCR text, may be able to identify and correct the errors using its
prior knowledge of valid Arabic.

%----------------------------------------------------------------------------
\section{The Open-Source vs.\ Closed-Source Gap}
\label{sec:intro_gap}
%----------------------------------------------------------------------------

The landscape of Arabic OCR is divided between commercial closed-source
systems and open-source alternatives. Closed-source vision--language models
(VLMs) --- including GPT-4V \cite{brown2020language}, Gemini, and similar
systems --- handle Arabic document recognition with high accuracy because
they were trained on massive multimodal corpora that include Arabic text in
diverse scripts, fonts, and document layouts. These systems are not publicly
auditable, cannot be deployed in offline or sensitive environments, and charge
per-query fees that are prohibitive for large archival digitisation projects.

Open-source OCR engines such as Qaari\footnote{Qaari is an open-source
Arabic OCR system developed for modern printed Arabic text. It is the
primary OCR engine evaluated in this thesis.} offer accessibility,
reproducibility, and deployability without network dependencies. However,
they fall behind closed-source alternatives on error rate, particularly on
difficult fonts, handwritten text, and documents with low scan quality.
The gap is meaningful: a 1--2\% CER difference translates to dozens of
errors per page in a long document, and a 10\% CER difference renders a
document largely unreadable to automated downstream systems.

The gap between open-source and closed-source systems exists at multiple
levels. At the recognition level, closed-source VLMs benefit from training
on billions of image--text pairs that include diverse Arabic document types,
while open-source engines are trained on smaller curated sets. At the
language-model level, closed-source systems incorporate large-scale language
models as priors that can use context to resolve visual ambiguity, while
open-source OCR engines rely on smaller, task-specific language components.
At the post-processing level, closed-source systems often include proprietary
correction layers that are not available to open-source users.

The question motivating this thesis is whether this gap can be narrowed
\emph{without changing the OCR engine itself}. Replacing an open-source
OCR engine with a closed-source one is not always feasible: cost, data
privacy, offline operation requirements, and the need for reproducibility
in academic research all create barriers to adopting closed-source solutions.
But adding a post-processing correction stage --- a language model that
receives the OCR output as text and corrects it before downstream use ---
is a modular intervention that leaves the OCR pipeline intact. If a language
model can reliably correct the systematic errors that open-source Arabic OCR
engines make, the effective accuracy of the system would improve without
requiring a new recognition model.

%----------------------------------------------------------------------------
\section{Post-OCR Correction as a Text-to-Text Task}
\label{sec:intro_postorc}
%----------------------------------------------------------------------------

Post-OCR correction treats the OCR output as noisy text and uses a language
model to produce a denoised version. The formulation is simple: given a
sequence of characters or words output by the OCR engine, produce the most
likely corresponding sequence in the original document. This is a
text-to-text task --- no images are required after the OCR stage --- and can
be approached with any sequence-to-sequence model.

The decoupling of the vision stage (OCR) from the language stage (correction)
has several practical advantages. It is engine-agnostic: the same correction
system can be applied to output from any OCR engine without modification.
It is retroactive: it can improve existing digitised archives without
re-scanning or re-running OCR, making it useful for large collections where
re-processing is impractical. It is modular: improvements to the OCR engine
and improvements to the corrector can be developed independently; adopting a
better correction model does not require retraining the recognition system,
and vice versa. And it is accessible: language models capable of Arabic text
processing are available as open-source checkpoints that can be run on
commodity hardware with GPU access, without proprietary APIs.

The rise of large language models (LLMs) trained on massive multilingual
corpora opens a new avenue for post-OCR correction. Unlike classical
correction approaches that require a specialised language model trained
on OCR error pairs \cite{tong1996statistical}, a general-purpose
instruction-tuned LLM can be directed to perform correction tasks through
a natural-language prompt. This \emph{prompt engineering} approach requires
no task-specific training, no labelled correction pairs, and no model
fine-tuning. The correction capability emerges from the model's pre-trained
knowledge of the target language.

However, prompt engineering is not a free lunch. The model's language priors
reflect the text it was trained on, which may not match the specific error
patterns of the OCR engine under evaluation. The model may over-correct,
rewriting text that is already correct; it may under-correct, leaving errors
that it cannot resolve from text context alone; or it may introduce new
errors by substituting plausible but incorrect words. Understanding when and
why these failure modes occur --- and whether injecting additional knowledge
can mitigate them --- is the core empirical question of this thesis.

%----------------------------------------------------------------------------
\section{Motivation for a Knowledge-Augmented Approach}
\label{sec:intro_motivation}
%----------------------------------------------------------------------------

A zero-shot LLM corrector operates on the model's general prior knowledge
of Arabic text. It knows that certain character sequences are rare or
non-words in Arabic, that certain word forms follow morphological patterns,
and that adjacent words should cohere semantically and syntactically. These
priors are powerful for catching clear errors but insufficient for catching
the systematic visual confusions that OCR engines produce --- confusions that
may yield perfectly plausible Arabic words that happen to be contextually
wrong.

Knowledge augmentation is the practice of injecting structured information
about the correction task into the model's prompt, supplementing its
generic language prior with task-specific guidance. In the context of
Arabic OCR correction, several knowledge sources are plausible:

\begin{itemize}

  \item \textbf{Character confusion matrices} derived from Phase~1 analysis
  can tell the model which specific character pairs the OCR engine tends to
  confuse. If the engine frequently outputs \RL{ن} where the ground truth is
  \RL{ب} on a particular font, injecting this information may help the model
  prioritise corrections of that specific substitution.

  \item \textbf{Self-reflective insights} derived from the model's own
  training-split correction behaviour capture the model's systematic blind
  spots: errors it consistently fails to fix, error types it introduces when
  correcting, and samples it over-corrects. By feeding these statistics back
  as prompt context, the model is given the opportunity to compensate for its
  own biases.

  \item \textbf{Morphological post-processing} via CAMeL Tools can catch
  a specific class of LLM output failures: words that are morphologically
  invalid in Arabic (i.e., not recognised by a morphological analyser).
  When the LLM introduces such an invalid word from a valid OCR source,
  a revert to the OCR original is a conservative and reliable corrective action.

  \item \textbf{Retrieval-augmented examples} drawn from Phase~2
  training-split corrections provide per-sample adaptive context.
  A BM25 index built on character n-grams of training OCR texts retrieves
  the $k$ most similar (OCR, ground-truth) pairs for each validation sample.
  These are injected into the prompt as concrete correction exemplars,
  tailoring the prompt to the specific error pattern of the current input.
  Unlike the global statistics of Phases~3 and~4, retrieved examples adapt
  sample-by-sample to the input at hand.

\end{itemize}

Each of these sources addresses a different aspect of the correction
problem, and they may interact in complex ways when combined. The central
question is not simply whether knowledge augmentation helps, but \emph{which}
knowledge type helps, \emph{when} it helps (for which fonts and error
profiles), and \emph{how much} each source contributes in an ablation study.

%----------------------------------------------------------------------------
\section{Research Questions}
\label{sec:intro_rq}
%----------------------------------------------------------------------------

This thesis investigates the following research questions. They are ordered
to reflect the experimental sequence: characterisation before correction,
isolated experiments before combinations.

\textbf{RQ1 --- Error Characterisation:} What are the characteristics of
errors produced by Qaari on Arabic typewritten and handwritten text? How do
error profiles differ across fonts and script types, and what does the error
taxonomy reveal about which correction strategies are most likely to succeed?

\textbf{RQ2 --- Zero-Shot Correction Capability:} Can a zero-shot
instruction-tuned LLM correct Arabic OCR errors without any task-specific
guidance? How does correction effectiveness vary across fonts with different
baseline error rates, and what is the nature and frequency of over-correction?

\textbf{RQ3 --- Confusion-Matrix Injection:} Does injecting the OCR engine's
known character confusion patterns (derived from Phase~1 analysis and
cross-referenced with LLM failure cases) into the system prompt improve
correction beyond zero-shot? Does the effect vary by font?

\textbf{RQ4 --- Self-Reflective Prompting:} Does feeding the model a
statistical summary of its own training-split correction failures --- fix
rates and introduction rates per error type, together with concrete
word-level error-correction pairs --- improve its correction quality on a
held-out validation set?

\textbf{RQ5 --- Morphological Post-Processing:} Does applying CAMeL Tools
morphological validation after zero-shot correction --- reverting
morphologically invalid LLM outputs to the valid OCR original, and reverting
known overcorrection patterns --- improve the quality of corrected text?

\textbf{RQ6 --- Strategy Combinations and Ablation:} What combination of
Phase~3 and Phase~4 strategies produces the best overall correction quality?
What is the marginal contribution of each component in ablation, and does
adding Phase~5 CAMeL post-processing provide further benefit on top of the
best prompt-only combination?

\textbf{RQ8 --- Retrieval-Augmented Correction:} Does per-sample retrieval
of similar OCR correction pairs from the training split --- via a BM25 index
over character n-grams --- improve correction quality beyond zero-shot
prompting? Does retrieving from the same domain and engine (training-split
corrections) overcome the domain-mismatch problem that limits external corpus
retrieval?

\textbf{RQ7 (overarching):} Does correction effectiveness depend primarily
on the baseline OCR error profile of the input, or can knowledge augmentation
overcome the over-correction problem observed on low-error fonts?

RQ1 is addressed in Chapter~\ref{ch_results}, Section~\ref{sec:res_phase1},
using Phase~1 outputs. RQ2--RQ5 are each addressed by a corresponding
isolated experiment (Phases~2--5), all reporting on the validation split.
RQ6 is addressed by the combination and ablation study (Phase~6). RQ8 is
addressed by the retrieval-augmented experiment (Phase~8). RQ7 is
addressed by synthesising findings across Phases~2--6 and~8 in
Chapter~\ref{ch_conc}.

%----------------------------------------------------------------------------
\section{Contributions}
\label{sec:intro_contributions}
%----------------------------------------------------------------------------

This thesis makes the following contributions.

\begin{enumerate}

  \item \textbf{A systematic multi-dataset evaluation framework for Arabic
  post-OCR correction.} Six prompt-engineering strategies are evaluated
  across nine datasets (eight PATS-A01 typewritten fonts and KHATT
  handwritten text) using aligned train/validation splits, each changing
  exactly one variable relative to the zero-shot baseline. This isolation
  design enables direct attribution of observed CER changes to individual
  knowledge sources, and the multi-font evaluation reveals font-dependent
  effects invisible in single-dataset studies. To the best of the author's
  knowledge, this is the first study to evaluate LLM prompt strategies for
  Arabic post-OCR correction simultaneously across this many font families
  under a controlled isolation design.

  \item \textbf{A three-stage export--infer--analyse pipeline.} The
  experimental infrastructure separates local data preparation from GPU
  inference from result analysis. Inference is executed on Kaggle (free GPU
  quota) via a unified \texttt{scripts/infer.py} script that handles
  resumption, HuggingFace synchronisation, and prompt versioning. The
  pipeline is fully modular: each phase adds one new \texttt{PromptBuilder}
  method and one pipeline script, with the inference and analysis layers
  unchanged. This design enables full reproducibility without a local GPU.

  \item \textbf{A novel self-reflective prompting strategy for OCR
  correction.} Phase~4 derives correction guidance from the model's own
  training-split error statistics --- quantifying fix rates and error
  introduction rates per error type, converting them into Arabic-language
  diagnostic sentences, and injecting both the statistical summary and
  concrete word-level OCR error pairs into the system prompt for
  validation-split inference. This strategy is, to the author's knowledge,
  the first application of self-diagnostic statistics as prompt augmentation
  for Arabic OCR correction. It achieves the strongest isolated result of any
  strategy tested: $-$0.7~pp CER on PATS-A01 relative to zero-shot.

  \item \textbf{An aligned multi-font evaluation design with quantified
  font effects.} All eight PATS-A01 fonts share 2,766 text lines drawn from
  the same underlying content, with an aligned 85/15 split (seed 42)
  ensuring identical 415-sample validation sets across fonts. This controlled
  design isolates font as the sole variable, enabling principled cross-font
  comparisons of both OCR quality and LLM correction effectiveness. The
  resulting per-font analysis quantifies the span of OCR error rates
  (4.3\%--24.3\% CER across fonts) and shows that font choice has a larger
  effect on effective accuracy than any correction strategy tested.

  \item \textbf{An empirical characterisation of the over-correction
  threshold.} Zero-shot LLM correction effectiveness depends strongly on the
  baseline error rate. On fonts with CER below approximately 8\%, the LLM
  introduces more errors than it fixes; on fonts with CER above this
  threshold, the net effect is positive. This font-dependent over-correction
  threshold is quantified across all eight typewritten fonts and KHATT,
  providing actionable guidance for when LLM correction should and should not
  be applied.

  \item \textbf{A comprehensive error taxonomy for Qaari across multiple
  fonts.} Phase~1 produces per-font character-level confusion matrices,
  error taxonomies with eight categories, and morphological analyses via
  CAMeL Tools, revealing how error profiles shift across fonts --- from
  dot-confusion-dominated errors on clean fonts to more diverse distributions
  on visually complex typefaces. The morphological analysis establishes that
  71\% of OCR errors are valid-but-wrong (real Arabic words in the wrong
  context), with important implications for Phase~5.

  \item \textbf{An ablation study of strategy combinations.} Phase~6 tests
  four combinations (\texttt{conf\_only}, \texttt{self\_only},
  \texttt{conf\_self}, \texttt{best\_camel}) and conducts a component-removal
  ablation to quantify each strategy's marginal contribution. The best
  combination (\texttt{conf\_self}) achieves 11.2\% CER on PATS-A01
  (21.1\% relative improvement over OCR baseline) and 42.4\% on KHATT
  (5.0\% relative improvement).

  \item \textbf{A retrieval-augmented correction strategy using same-domain
  training examples.} Phase~8 builds a BM25 index over character n-grams
  of Phase~2 training corrections and, for each validation sample, retrieves
  the $k$ most similar (OCR, ground-truth) pairs to inject as per-sample
  few-shot context. This design sidesteps the domain-mismatch problem that
  caused external corpus retrieval (OpenITI) to degrade performance, by
  restricting the retrieval pool to the same OCR engine and text domain as
  the evaluation data. To the author's knowledge, this is the first
  application of BM25 character-n-gram retrieval to Arabic post-OCR
  correction, providing an adaptive, per-sample complement to the global
  statistical patterns of Phases~3 and~4.

  \item \textbf{Open-source code and reproducible configuration.}
  All pipelines, prompt templates, knowledge base loaders, and evaluation
  scripts are publicly available. Every experimental result is reproducible
  from the repository using the provided configuration files and inference
  notebooks. The \texttt{MockCorrector} backend enables local development
  and testing of all phases without GPU access.

\end{enumerate}

%----------------------------------------------------------------------------
\section{Scope and Boundaries}
\label{sec:intro_scope}
%----------------------------------------------------------------------------

Several design decisions define the scope of this thesis and bound the
generalisability of its findings.

\textbf{Single OCR engine.} All experiments use Qaari as the OCR engine.
Qaari's specific error patterns --- particularly its character confusion
profile across fonts --- shape the results of all subsequent phases,
including which confusion patterns are injected in Phase~3 and which
self-reflective insights are derived in Phase~4. Results may differ for
other OCR engines (e.g., Kraken, EasyOCR, commercial engines) with
different error profiles.

\textbf{Single LLM.} All correction experiments use
Qwen3-4B-Instruct-2507. Larger models (7B, 13B, 70B) may have stronger
Arabic language priors and different over-correction behaviour. The choice
of Qwen3-4B was motivated by its multilingual capability, its fit within
the free-tier GPU compute available on Kaggle (16~GB VRAM), and its strong
Arabic performance relative to its parameter count.

\textbf{Text-only correction.} All phases treat correction as a pure
text-to-text task. The document image is not used after the OCR stage. This
means that visually ambiguous cases --- where the correct character can only
be determined from the image --- are inaccessible to the corrector. The
results on KHATT handwritten text partly reflect this limitation.

\textbf{Modern Standard Arabic.} PATS-A01 documents are printed in MSA.
KHATT contains handwritten MSA. Results on dialectal Arabic, Classical
Arabic, or code-switched text (Arabic--English mixing) may differ.

\textbf{Prompt engineering without fine-tuning.} All strategies are
zero-shot or few-shot prompt-engineering approaches. Fine-tuning the LLM
on OCR error pairs is outside the scope of this thesis. Fine-tuned models
might outperform all prompt strategies tested but require labelled training
data and substantially greater compute.

%----------------------------------------------------------------------------
\section{Thesis Organisation}
\label{sec:intro_organisation}
%----------------------------------------------------------------------------

The remainder of this thesis is organised as follows.

\textbf{Chapter~\ref{ch_background} --- Background and Literature Review}
builds the necessary foundations. It covers the Arabic language and script
in depth, the evolution of OCR technology from early rule-based systems
through neural approaches, the development of large language models from
the transformer architecture through instruction tuning and alignment, and
prompt engineering strategies relevant to this study (zero-shot, chain-of-
thought, self-reflective prompting). It surveys related work on post-OCR
correction (classical and neural), Arabic NLP resources, and morphological
analysis via CAMeL Tools. A reader familiar with any of these topics may
skip the corresponding subsection.

\textbf{Chapter~\ref{ch_system} --- System Design and Datasets}
describes the complete experimental infrastructure. It introduces the
PATS-A01 and KHATT datasets in detail, explains the aligned multi-font
train/validation split design and its rationale, describes the three-stage
export--infer--analyse pipeline, defines the text normalisation and
CER/WER evaluation protocol, documents each knowledge base component
(confusion matrix loader and LLM insights loader), and presents the
\texttt{PromptBuilder} class hierarchy and all prompt variants used across
phases.

\textbf{Chapter~\ref{ch_methodology} --- Experimental Methodology}
describes each experimental phase as a self-contained experiment with a
stated hypothesis, step-by-step method, variables changed vs.\ Phase~2,
output files, and the research question answered. Phases~1 through~5 are
covered as isolated experiments, followed by Phase~6 (combinations and
ablation) and Phase~8 (retrieval-augmented generation), with particular
attention to the isolation principle that ensures each augmentation condition
can be attributed to a single knowledge source.

\textbf{Chapter~\ref{ch_results} --- Results and Analysis}
reports all numerical results across the nine evaluation datasets.
Per-font Phase~1 baselines establish the OCR error landscape. Phase~2
zero-shot results reveal font-dependent correction effectiveness. Phases~3--5
quantify the impact of each knowledge source, and Phase~6 tests combinations
with ablation. Every result is presented in a table and interpreted in prose,
with statistical significance reported via paired t-tests and Cohen's~$d$
effect sizes.

\textbf{Chapter~\ref{ch_conc} --- Discussion and Conclusion}
interprets the findings, develops the font-dependent over-correction
threshold as a unified theoretical argument, revisits each research question
in the light of the results, discusses the conditions under which the findings
would likely differ, states practical system design recommendations, enumerates
limitations, and charts future research directions.

The appendices provide supporting material: Appendix~A contains the full
prompt templates for all phases; Appendix~B contains the Phase~4
Arabic-language diagnostic text generated from training-split analysis;
Appendix~C contains the relevant sections of the configuration file; and
Appendix~D contains representative tables of OCR error examples from Phase~1.
