\setstretch{1.5}

This appendix contains the complete system prompt templates used in each
experimental phase. All prompts are written in Arabic, as described in
Section~\ref{sec:promptbuilder}. English translations are provided
for accessibility. The user message in every phase is the raw OCR text
to be corrected; only the system messages differ across phases.
Phase~5 (CAMeL post-processing) uses the same zero-shot prompt as Phase~2
and applies its morphological revert logic to the model output rather than
changing the prompt.

%----------------------------------------------------------------------------
\section*{A.1 \quad Phase 2 --- Zero-Shot Prompt (v1, Aggressive)}
%----------------------------------------------------------------------------

\textbf{Version tag:} \texttt{p2v1} \quad
\textbf{Source:} \texttt{PromptBuilder.ZERO\_SHOT\_SYSTEM\_V1}

\noindent
This was the initial prompt design, adopted in early pilot runs.
It makes no distinction between texts that are already correct and those that
contain errors, and simply instructs the model to perform correction.
Pilot results showed that this aggressive framing caused the model to rewrite
clean text unnecessarily, inflating CER on low-error fonts.
It was superseded by the conservative v2 variant for all reported experiments.

\begin{quote}\itshape
  \RL{أنت مصحح نصوص عربية متخصص.}
  \RL{مهمتك تصحيح أخطاء التعرف الضوئي على الحروف (OCR) في النص العربي.}
  \RL{أعد فقط النص المصحح دون أي شرح أو تعليق إضافي.}
\end{quote}

\textbf{English translation:}
\begin{quote}
  You are a specialised Arabic text corrector.
  Your task is to correct optical character recognition (OCR) errors in
  Arabic text.
  Return only the corrected text with no explanation or additional commentary.
\end{quote}

%----------------------------------------------------------------------------
\section*{A.2 \quad Phase 2 --- Zero-Shot Prompt (v2, Conservative)}
%----------------------------------------------------------------------------

\textbf{Version tag:} \texttt{p2v2} \quad
\textbf{Source:} \texttt{PromptBuilder.ZERO\_SHOT\_SYSTEM\_V2}

\noindent
This is the primary prompt used in all reported experiments for Phase~2 and
Phase~5. The conservative framing explicitly tells the model that the input
may already be correct and that it should return it unchanged in that case.
This is critical for low-error fonts such as Akhbar, where the majority of
tokens require no change.

\begin{quote}\itshape
  \RL{أنت مدقق لغوي للنصوص العربية.}
  \RL{النص التالي هو مخرجات نظام OCR وقد يكون صحيحاً تماماً أو يحتوي على أخطاء قليلة.}
  \RL{مهمتك: إذا كان النص صحيحاً، أعده كما هو دون أي تغيير.}
  \RL{إذا وجدت خطأً واضحاً فقط، صحِّحه. لا تغيِّر ما هو صحيح.}
  \RL{أعد فقط النص دون أي شرح.}
\end{quote}

\textbf{English translation:}
\begin{quote}
  You are an Arabic text proofreader.
  The following text is the output of an OCR system and may be completely
  correct or may contain a few errors.
  Your task: if the text is correct, return it exactly as is.
  If you find only a clear error, correct it. Do not change what is correct.
  Return only the text with no explanation.
\end{quote}

%----------------------------------------------------------------------------
\section*{A.3 \quad Phase 3 --- OCR-Aware Prompt (p3v1)}
%----------------------------------------------------------------------------

\textbf{Version tag:} \texttt{p3v1} \quad
\textbf{Source:} \texttt{PromptBuilder.build\_ocr\_aware()}

\noindent
The placeholder \texttt{\{confusion\_context\}} is replaced at runtime with
the top-10 confusion pairs formatted by \texttt{ConfusionMatrixLoader}.
Each pair is a bullet item naming a ground-truth character and the character
Qaari most commonly substitutes in its place.

\begin{quote}\itshape
  \RL{أنت مصحح نصوص عربية متخصص في تصحيح مخرجات نظام التعرف الضوئي Qaari على الحروف.}
  \RL{فيما يلي أكثر الأخطاء شيوعاً التي يرتكبها هذا النظام مع الحروف العربية:}

  \texttt{\{confusion\_context\}}

  \RL{صحِّح النص التالي مع الانتباه بشكل خاص لهذه الأخطاء.}
  \RL{أعد فقط النص المصحح دون أي شرح أو تعليق إضافي.}
\end{quote}

\textbf{English translation:}
\begin{quote}
  You are an Arabic text corrector specialised in correcting the output of the
  Qaari optical character recognition system.
  The following are the most common errors made by this system with Arabic
  characters:

  \{confusion\_context\}

  Correct the following text paying special attention to these errors.
  Return only the corrected text with no explanation or additional commentary.
\end{quote}

%----------------------------------------------------------------------------
\section*{A.4 \quad Phase 4 --- Self-Reflective Prompt (p4v3)}
%----------------------------------------------------------------------------

\textbf{Version tag:} \texttt{p4v3} \quad
\textbf{Source:} \texttt{PromptBuilder.build\_self\_reflective()}

\noindent
The placeholder \texttt{\{insights\_context\}} is replaced at runtime with
the Arabic diagnostic text from \texttt{LLMInsightsLoader}
(see Appendix~B for example generated insight text).
The insights are derived from Phase~2 corrections on the training split and
include per-category fix rates, weakness categories, and over-correction
warnings. Concrete word-level error-correction pairs from the training split
(\texttt{word\_pairs\_llm\_failures.txt}) are also injected alongside
the statistical diagnostics.

\begin{quote}\itshape
  \RL{أنت مصحح نصوص عربية متخصص. بناءً على تحليل أخطائك السابقة في تصحيح نصوص عربية مشابهة،}
  \RL{فيما يلي ملاحظات مهمة لتحسين أدائك:}

  \texttt{\{insights\_context\}}

  \RL{صحِّح النص التالي مع مراعاة هذه الملاحظات.}
  \RL{أعد فقط النص المصحح دون أي شرح أو تعليق إضافي.}
\end{quote}

\textbf{English translation:}
\begin{quote}
  You are a specialised Arabic text corrector. Based on an analysis of your
  previous errors correcting similar Arabic texts, here are important notes
  for improving your performance:

  \{insights\_context\}

  Correct the following text taking these notes into account.
  Return only the corrected text with no explanation or additional commentary.
\end{quote}

%----------------------------------------------------------------------------
\section*{A.5 \quad Phase 6 --- Combined Prompt (p6v1)}
%----------------------------------------------------------------------------

\textbf{Version tag:} \texttt{p6v1} \quad
\textbf{Source:} \texttt{PromptBuilder.build\_combined()}

\noindent
The combined prompt assembles only non-empty context sections in a fixed
order: (1)~confusion context, (2)~self-reflective insights. Below is the
full template with both sections present, corresponding to the
\texttt{conf\_self} combination. When only one section is requested
(\texttt{conf\_only} or \texttt{self\_only}), the other placeholder
and its section heading are omitted entirely.

\begin{quote}\itshape
  \RL{أنت مصحح نصوص عربية متخصص. استخدم المعلومات التالية لتصحيح أخطاء OCR:}

  \RL{أولاً: الأخطاء الشائعة في هذا النظام:}
  \texttt{\{confusion\_context\}}

  \RL{ثانياً: ملاحظات تحسين الأداء الذاتي:}
  \texttt{\{insights\_context\}}

  \RL{صحِّح النص التالي. أعد فقط النص المصحح دون أي شرح أو تعليق.}
\end{quote}

\textbf{English translation:}
\begin{quote}
  You are a specialised Arabic text corrector. Use the following information
  to correct OCR errors:

  First: common errors in this system:
  \{confusion\_context\}

  Second: self-improvement performance notes:
  \{insights\_context\}

  Correct the following text. Return only the corrected text with no
  explanation or commentary.
\end{quote}

\noindent
For the \texttt{best\_camel} combination, the same \texttt{conf\_self}
prompt is used, with the CAMeL morphological revert logic
(\texttt{WordValidator.validate\_correction()}) applied as a post-processing
step to the model output. No additional placeholder is added to the system
message; the revert is entirely post-hoc.
