\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{url}
\usepackage{makecell}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Does More Context Help? Evaluating Knowledge-Augmented Prompt Strategies
for Arabic Post-OCR Correction with a Lightweight LLM}

\author{\IEEEauthorblockN{Author Name}
\IEEEauthorblockA{\textit{Department of Computer Science} \\
\textit{University Name}\\
City, Country \\
email@university.edu}
}

\maketitle

\begin{abstract}
Arabic optical character recognition (OCR) systems produce systematic
character-level errors---dot confusions, hamza substitutions, taa marbuta
swaps---that degrade downstream text processing. Post-OCR correction using
large language models (LLMs) is a natural remedy, but it is unclear how much
additional context actually helps, and whether findings generalize across
font families and script types. This paper reports a controlled experiment
in which five prompt-engineering strategies are evaluated across nine
datasets---eight typewritten fonts from PATS-A01 and one handwritten set from
KHATT---using Qwen3-4B-Instruct-2507 as the correction model. The strategies
tested are: zero-shot correction (baseline), confusion-matrix injection,
self-reflective prompting with training-split error analysis, morphological
post-processing via CAMeL Tools, and optimal strategy combinations with
ablation. Results reveal that correction effectiveness is strongly
font-dependent: zero-shot LLM correction reduces CER by up to 81\% on
high-error typewritten fonts but can increase errors on fonts where Qaari
already performs well. On handwritten text (KHATT), where baseline OCR errors
exceed 44\% CER, the LLM provides marginal improvement. The best combination
(confusion-matrix injection + self-reflective prompting) achieves 11.2\% CER
on PATS-A01 (down from 14.2\% OCR baseline) and 42.4\% on KHATT (down from
44.6\%). These findings suggest that the optimal correction strategy depends
on the baseline error profile, and that a one-size-fits-all approach to prompt
augmentation is unlikely to succeed.
\end{abstract}

\begin{IEEEkeywords}
Arabic OCR, post-OCR correction, large language models, prompt engineering,
Qwen3, PATS-A01, KHATT
\end{IEEEkeywords}

% -------------------------------------------------------------------------
\section{Introduction}
% -------------------------------------------------------------------------

Arabic optical character recognition occupies a uniquely difficult position
in the document analysis landscape. The script is inherently cursive, with
letter forms that vary by position within the word (isolated, initial, medial,
final); dots serve as the sole phonemic and semantic distinguisher for entire
character groups (\textit{ba}/\textit{ta}/\textit{tha}/\textit{ya}/\textit{nun}
share an identical skeletal form); and the optional nature of diacritics
introduces a pervasive ambiguity that complicates both recognition and
evaluation. These properties conspire to produce systematic, structured errors
in open-source OCR engines such as Qaari~\cite{qaari}---errors that propagate
unchecked into downstream pipelines for information retrieval, machine
translation, and text-to-speech synthesis.

Post-OCR correction---treating recognition output as noisy text and applying a
language model as a denoiser---has attracted renewed attention since
instruction-tuned large language models (LLMs) became widely available. The
paradigm is appealing: it requires no task-specific training data, operates in a
single forward pass, and leverages the model's implicit prior over valid
linguistic sequences. Yet two questions remain inadequately addressed in the
literature. First, \emph{does injecting structured knowledge about the OCR
engine's failure modes into the prompt yield measurable gains over zero-shot
correction?} Second, and more fundamentally, \emph{does correction effectiveness
generalise across font families and script types, or is it an artefact of the
specific evaluation conditions?}

To answer these questions, we design a controlled multi-dataset experiment.
Starting from a zero-shot correction baseline, we progressively augment the
prompt with three orthogonal knowledge sources: (i)~a failure-filtered character
confusion matrix extracted from the OCR engine's systematic substitution patterns,
(ii)~self-reflective diagnostic text derived from the model's own training-split
correction failures---quantifying its fix rate, introduction rate, and concrete
word-level error pairs---and (iii)~morphological post-processing via CAMeL
Tools~\cite{obeid2020} to revert corrections that produce morphologically invalid
Arabic. Each augmentation is first evaluated in isolation (changing exactly one
variable relative to zero-shot), then in combination with ablation analysis. All
nine datasets---eight typewritten fonts from PATS-A01~\cite{pats} and one
handwritten corpus from KHATT~\cite{khatt}---are evaluated under an aligned
train/validation split that holds text content constant across fonts, isolating
font as the sole confound.

The central finding is that correction effectiveness is governed by a
\emph{font-dependent over-correction threshold}: a baseline CER below which the
LLM's tendency to rewrite correct tokens outweighs its error-fixing capability,
and above which the net effect is beneficial. On typewritten fonts with
moderate-to-high error rates (10--25\% CER), zero-shot correction yields
substantial reductions (up to 81\% relative on Traditional font). On the
lowest-error font (Akhbar, 4.3\% CER), every strategy tested increases errors.
On handwritten text (KHATT, 44.6\% CER), where 37\% of errors are segmentation
failures inaccessible to text-only correction, improvement is marginal. Knowledge
augmentation provides further incremental gains above zero-shot, but its benefit
is also font-dependent, and the most impactful strategy
(self-reflective prompting) achieves its advantage partly by explicitly warning
the model against over-correction.

The contributions of this work are:
\begin{itemize}
\item A systematic, controlled comparison of five prompt-engineering strategies
      for Arabic post-OCR correction across nine datasets spanning an order-of-magnitude
      CER range (4.3--44.6\%), revealing font-dependent behaviour that
      single-dataset studies structurally cannot detect.
\item A formal characterisation of the \emph{over-correction threshold}: the
      baseline CER below which LLM correction is net harmful, estimated
      empirically at approximately 6--7\% for Qwen3-4B on Qaari output.
\item A novel self-reflective prompting strategy that injects the model's own
      training-split error statistics and concrete word-level failure pairs into
      the system prompt, achieving the strongest isolated improvement ($-$0.7~pp
      CER) and uniquely improving the lowest-error font.
\item An ablation study establishing that confusion-matrix injection and
      self-reflective prompting address orthogonal failure modes and combine
      near-additively, with the self-reflective component contributing
      approximately twice the marginal CER reduction.
\end{itemize}

% -------------------------------------------------------------------------
\section{Related Work}
% -------------------------------------------------------------------------

\subsection{Arabic OCR and Its Error Patterns}
Arabic OCR has been studied for over three decades~\cite{arabic_ocr_survey}.
The core difficulties are structural: the cursive baseline couples segmentation
with recognition (a segmentation error deterministically produces a recognition
error), diacritics are contextually optional yet semantically significant, and
several letter groups (ba/ta/tha/ya/nun; fa/qaf; dal/dhal; ra/zay) share
identical skeletal forms differing only in the number and placement of dots.
These dot-group confusions are the dominant error type in modern Arabic OCR
engines and are the primary target of the confusion-matrix augmentation
evaluated in this work. Qaari, the open-source engine used throughout this
study, is a vision--language model (VLM) fine-tuned from Qwen2-VL-2B for
Arabic document recognition~\cite{qaari}; it inherits the strengths of its
base architecture while exhibiting systematic dot-confusion patterns that are
font-dependent and quantifiable.

\subsection{Post-OCR Correction}
Post-OCR correction has evolved through three paradigms.
\textit{Rule-based and statistical} methods employed lexicon lookup, n-gram
language models, and noisy-channel formulations that model OCR errors as a
probabilistic channel between the true text and the observed
output~\cite{afli2016}. These methods are effective for high-frequency
error patterns but require manual rule curation and struggle with unseen
error types.
\textit{Neural sequence-to-sequence} approaches treat correction as a
translation task, training encoder--decoder models on parallel (OCR, GT)
corpora~\cite{dong2018,nguyen2021}. While powerful, they require substantial
training data and are tightly coupled to the specific OCR engine's error
distribution.
\textit{LLM-based} methods, the paradigm explored here, leverage pre-trained
instruction-tuned models to correct OCR output via natural-language prompts
without task-specific training. Recent work demonstrates that LLMs can
outperform specialised models on English post-OCR correction~\cite{llm_ocr_correction},
but Arabic-specific evaluations remain limited to single-dataset studies using
BERT-class masked language models~\cite{arabic_bert_ocr}. No prior work
systematically evaluates multiple prompt augmentation strategies across
multiple Arabic fonts under a controlled isolation design.

\subsection{Knowledge-Augmented Prompting}
The injection of structured knowledge into LLM prompts has demonstrated
consistent gains across NLP tasks. Few-shot exemplars activate task-specific
priors without parameter updates~\cite{brown2020}; chain-of-thought prompting
decomposes complex reasoning into verifiable intermediate
steps~\cite{wei2022}; retrieval-augmented generation (RAG) grounds model
outputs in external evidence to reduce hallucination~\cite{lewis2020}. Most
relevant to this work, Reflexion~\cite{shinn2023} and related self-critique
frameworks feed an agent its own prior failures as verbal reinforcement,
enabling behavioural adaptation without gradient updates. Our Phase~4
self-reflective strategy adapts this paradigm to a non-agentic, single-turn
correction task: the model receives a statistical summary of its own
training-split errors as system-prompt context, including quantified fix/introduction
rates and concrete word-level failure pairs. To our knowledge, this is the first
application of self-diagnostic prompting to post-OCR correction in any language.

\subsection{Arabic NLP Resources}
CAMeL Tools~\cite{obeid2020} provides morphological analysis, disambiguation,
and tokenisation for Modern Standard Arabic using the CALIMA Star
analyser~\cite{habash2010}. We use its morphological component to validate
correction outputs (Phase~5) and to characterise the valid-but-wrong error
rate in Phase~1. The PATS-A01~\cite{pats} and KHATT~\cite{khatt} benchmarks
provide complementary evaluation data: PATS-A01 offers eight typewritten fonts
rendering identical text content, enabling controlled cross-font comparison;
KHATT provides real-world handwritten variation with substantially higher
error rates and qualitatively different error types (segmentation-dominated
rather than substitution-dominated).

% -------------------------------------------------------------------------
\section{Datasets and Experimental Setup}
% -------------------------------------------------------------------------

\subsection{Datasets}
All experiments use two complementary Arabic OCR benchmarks.

\textbf{PATS-A01}~\cite{pats} is a typewritten Arabic dataset covering eight
font families: Akhbar, Andalus, Arial, Naskh, Simplified, Tahoma, Thuluth,
and Traditional. Each font subset contains synthetic document images with
known ground-truth text rendered in a different typeface. We use an 85/15
train/validation split (seed=42), yielding 415 validation samples per font
from a shared pool of 2{,}766 text lines. Crucially, the split is
\textit{aligned} across fonts: every font's validation set contains the same
text content, so that differences in CER across fonts can be attributed
solely to the typeface, not the text. Training splits are used for
self-reflective analysis (Phase~4); validation splits are used for all
reported metrics.

\textbf{KHATT}~\cite{khatt} is a handwritten Arabic dataset with real-world
variation in writing style, pen type, and image quality. It provides a
significantly harder challenge than typewritten text: baseline OCR error rates
exceed 44\% CER, compared to single-digit CER for the best PATS-A01 font.

Table~\ref{tab:datasets} summarizes the evaluation datasets.

\begin{table}[t]
\caption{Evaluation datasets. PATS-A01 fonts share the same text content
rendered in different typefaces. KHATT is a separate handwritten corpus.}
\label{tab:datasets}
\begin{center}
\small
\begin{tabular}{llr}
\toprule
\textbf{Dataset} & \textbf{Type} & \textbf{Val. samples} \\
\midrule
PATS-A01-Akhbar       & Typewritten & 415 \\
PATS-A01-Andalus      & Typewritten & 415 \\
PATS-A01-Arial        & Typewritten & 415 \\
PATS-A01-Naskh        & Typewritten & 415 \\
PATS-A01-Simplified   & Typewritten & 415 \\
PATS-A01-Tahoma       & Typewritten & 415 \\
PATS-A01-Thuluth      & Typewritten & 415 \\
PATS-A01-Traditional  & Typewritten & 415 \\
KHATT-validation      & Handwritten & 1{,}215 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

Diacritics are stripped before CER and WER evaluation to avoid penalizing the
model for diacritic insertion or deletion, which is not the target of this
correction task. Both diacritics-stripped and raw metrics are computed; we
report the stripped variant throughout.\footnote{Results with diacritics
retained show the same relative patterns.}

\subsection{OCR Engine}
We use Qaari as the OCR engine. Qaari's output serves as both the baseline
for Phase~1 analysis and the input to all LLM correction phases.

\subsection{Correction Model}
Qwen3-4B-Instruct-2507~\cite{qwen3} is used as the LLM for all correction
phases. It is a 4-billion-parameter instruction-tuned model with strong
multilingual capabilities and native Arabic support. Inference was run on
a GPU-accelerated environment (Kaggle T4). The \texttt{enable\_thinking} flag
was set to \texttt{False} to suppress the chain-of-thought scratchpad, which
increases token cost without measurable benefit for this task.

\subsection{Evaluation Metrics}
Character Error Rate (CER) is the primary metric:
\begin{equation}
\text{CER} = \frac{S + D + I}{N},
\label{eq:cer}
\end{equation}
where $S$, $D$, $I$ are the number of substitutions, deletions, and insertions
at the character level, and $N$ is the total number of characters in the
reference. Word Error Rate (WER) is reported alongside CER for completeness.
Lower values indicate better performance.

\subsection{Pipeline Architecture}
Each experimental phase follows a three-stage pipeline: (1)~\textit{export},
which prepares the inference input JSONL from the local dataset;
(2)~\textit{infer}, which runs the LLM on a GPU-capable environment with
periodic checkpointing to HuggingFace Hub for fault tolerance; and
(3)~\textit{analyze}, which computes metrics and comparison statistics locally.
This separation allows the correction model to run on cloud GPUs while
evaluation happens on the local machine with access to ground-truth data.

% -------------------------------------------------------------------------
\section{Experimental Methodology}
% -------------------------------------------------------------------------

We organize the experiment into six phases. Phase~1 characterizes the OCR
errors. Phase~2 is the zero-shot LLM correction baseline. Phases~3, 4, and~5
each introduce exactly one form of additional knowledge, isolating the effect
of each augmentation relative to Phase~2. Phase~6 tests combinations of the
isolated strategies with ablation analysis.

\subsection{Phase 1: OCR Baseline and Error Taxonomy}

Before correcting anything, we characterize what there is to correct.
Phase~1 runs Qaari over all datasets and computes CER and WER against
ground truth. It also builds a character-level confusion matrix per dataset
and classifies errors into categories: dot confusion (ba/ta/tha/ya
misreads), hamza errors, taa marbuta/ha swaps, alef-maksura/ya swaps,
similar-shape pairs (ra/zay, dal/dhal), and other substitutions.

Morphological validity of OCR tokens is checked with CAMeL
Tools~\cite{obeid2020}. Words that fail morphological analysis are classed as
non-word errors; those that parse successfully but differ from the ground
truth are valid-but-wrong errors. This distinction is important because
non-word errors should be straightforward for an LLM to catch, whereas
valid-but-wrong errors require contextual understanding.

\subsection{Phase 2: Zero-Shot LLM Correction (Baseline)}

The zero-shot prompt instructs the model to correct OCR errors in the supplied
text and return only the corrected output without explanation. The system prompt
uses an XML-structured template specifying the task, nine correction rules
(preserve meaning, fix visual OCR errors only, do not add diacritics, etc.),
and few-shot format examples demonstrating the expected input--output structure.
This conservative formulation was chosen to reduce unnecessary rewrites, which
are a known failure mode for instruction-following models on text that is mostly
correct.

Phase~2 output serves as the comparison point for all subsequent phases. A
phase is considered beneficial only if it achieves lower CER than Phase~2.

\subsection{Phase 3: OCR-Aware Prompting (Confusion-Matrix Injection)}

Phase~3 extends the system prompt with a list of Qaari's most frequent
character confusions extracted from the Phase~1 confusion matrix. Specifically,
the top-10 character-pair confusions per dataset are formatted as an Arabic
bullet list and injected as a \texttt{<confusion\_patterns>} section before the
correction examples. To focus on actionable confusions, the list is
cross-referenced with the model's own failure cases from the Phase~2 training
split: only confusions that appear in samples the model failed to correct are
retained. The hypothesis is that telling the model which characters Qaari tends
to mix up---filtered to those the model itself has struggled with---should focus
its attention on the most impactful patterns.

\subsection{Phase 4: Self-Reflective Prompting}

Phase~4 feeds the model a statistical summary of its own past mistakes. To
generate this summary, Phase~2 is run on the training split, and the
corrections are compared to ground truth by an error analyzer that tracks the
fix rate and introduction rate per error type. The resulting artifacts---word
pairs from unfixed and newly introduced errors---are saved to disk and loaded
by Phase~4 at inference time. The insights are aggregated separately for
PATS-A01 and KHATT (reflecting their different error profiles) and formatted
as Arabic-language diagnostic notes injected in a \texttt{<self\_analysis>}
section.

Additionally, concrete word-level OCR error pairs extracted from the training
split are included, showing the model specific examples of
\textit{ocr\_word}~$\to$~\textit{gt\_word} corrections it should apply. This
combines statistical self-awareness with concrete exemplars from the model's
own failure cases, making Phase~4 a targeted self-corrective approach.

\subsection{Phase 5: CAMeL Morphological Post-Processing}

Unlike Phases~3 and~4, Phase~5 does not modify the prompt. Instead,
it applies a post-processing step after zero-shot correction: each word in the
LLM output is checked against the CAMeL Tools morphological
analyzer~\cite{obeid2020}. If a corrected word fails morphological analysis
\textit{and} the original OCR word passes it, the corrected word is reverted
to the OCR original. Additionally, words that appear in a known-overcorrection
list---derived from training-split analysis of cases where the LLM changed a
correct word to an incorrect one---are reverted regardless of morphological
status. The strategy is conservative by design: it only intervenes when the
LLM has produced a morphologically invalid word from a valid source, or
repeated a documented overcorrection pattern.

\subsection{Phase 6: Combinations and Ablation}

Phase~6 systematically tests combinations of the isolated strategies from
Phases~3, 4, and~5. Four combinations are evaluated:

\begin{itemize}
\item \textbf{conf\_only}: Phase~3 confusion-matrix injection alone.
\item \textbf{self\_only}: Phase~4 self-reflective prompting alone.
\item \textbf{conf\_self}: Phase~3 and Phase~4 combined (both prompt
      augmentations active simultaneously).
\item \textbf{best\_camel}: Phase~3 + Phase~4 with Phase~5 CAMeL
      post-processing applied on top.
\end{itemize}

Ablation analysis compares each combination to the Phase~2 zero-shot baseline,
and also measures the marginal contribution of each component by removing it
from the best-performing combination.

% -------------------------------------------------------------------------
\section{Results}
% -------------------------------------------------------------------------

\subsection{Phase 1: OCR Baseline Error Profile}

Table~\ref{tab:phase1} shows the raw OCR error rates across all nine
evaluation datasets (diacritics stripped).

\begin{table}[t]
\caption{Phase 1: Qaari OCR baseline CER and WER per dataset (diacritics
stripped). Fonts are ordered by increasing CER.}
\label{tab:phase1}
\begin{center}
\small
\begin{tabular}{lrr}
\toprule
\textbf{Dataset} & \textbf{CER (\%)} & \textbf{WER (\%)} \\
\midrule
PATS-A01-Akhbar       & 4.3  & 8.7  \\
PATS-A01-Simplified   & 7.9  & 14.6 \\
PATS-A01-Thuluth      & 12.1 & 23.1 \\
PATS-A01-Traditional  & 14.2 & 25.8 \\
PATS-A01-Arial        & 14.8 & 28.2 \\
PATS-A01-Andalus      & 15.3 & 29.4 \\
PATS-A01-Tahoma       & 20.6 & 38.1 \\
PATS-A01-Naskh        & 24.3 & 43.0 \\
\midrule
KHATT-validation      & 44.6 & 72.4 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The error rate varies substantially across fonts, spanning nearly an order of
magnitude (4.3\% to 24.3\% CER) within the PATS-A01 typewritten benchmark.
This variation is itself a finding: the choice of typeface has a larger effect
on OCR accuracy than any post-processing strategy we test. The PATS-A01
macro-average CER is 14.2\%. KHATT (44.6\% CER) presents a qualitatively
different challenge.

The error taxonomy reveals that PATS-A01 errors are dominated by dot
confusions and similar-shape substitutions, while KHATT errors are more
diverse, including segmentation failures, character merges, and splits that
are rare in typewritten text.

Morphological analysis shows that across all datasets, 71\% of OCR errors
are of the valid-but-wrong type---the OCR produces real Arabic words that
happen to be incorrect in context. Non-word errors (tokens that fail
morphological analysis) account for 29\% of total errors. This has important
implications for Phase~5, which can only reliably catch non-word errors.

\subsection{Phase 2: Zero-Shot LLM Correction}

Table~\ref{tab:phase2} shows the effect of zero-shot LLM correction across
all datasets.

\begin{table}[t]
\caption{Phase 2: Zero-shot LLM correction vs.\ OCR baseline (diacritics
stripped). $\Delta$CER is absolute change in percentage points; negative
indicates improvement.}
\label{tab:phase2}
\begin{center}
\small
\begin{tabular}{lrrr}
\toprule
\textbf{Dataset} & \textbf{OCR CER} & \textbf{LLM CER} & \textbf{$\Delta$CER} \\
\midrule
PATS-Akhbar       & 4.3  & 5.2  & +0.9 \\
PATS-Simplified   & 7.9  & 7.3  & -0.6 \\
PATS-Thuluth      & 12.1 & 11.2 & -0.9 \\
PATS-Traditional  & 14.2 & 2.7  & -11.5 \\
PATS-Arial        & 14.8 & 14.7 & -0.1 \\
PATS-Andalus      & 15.3 & 15.1 & -0.2 \\
PATS-Tahoma       & 20.6 & 20.5 & -0.1 \\
PATS-Naskh        & 24.3 & 21.5 & -2.8 \\
\midrule
KHATT-val         & 44.6 & 43.0 & -1.6 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The results reveal a strong font-dependent pattern. The PATS-A01 macro-average
CER drops from 14.2\% to 12.3\% (a 13.4\% relative improvement). The most
dramatic gain is on PATS-A01-Traditional, where zero-shot correction achieves
an 81.0\% relative CER reduction (14.2\% $\to$ 2.7\%). Naskh also benefits
substantially ($-$11.5\% relative). By contrast, on Akhbar---where Qaari
already achieves only 4.3\% CER---zero-shot correction \textit{increases}
errors ($+$0.9 pp), a phenomenon we term \textit{over-correction on
low-error input}: the model rewrites tokens that Qaari had rendered correctly.

On KHATT (handwritten), the improvement is marginal ($-$1.6 pp, 3.6\%
relative) despite the high baseline error rate. Handwritten OCR errors involve
character segmentation failures and stroke ambiguities that are harder for the
LLM to resolve from text alone without visual context.

\subsection{Phases 3--5: Isolated Knowledge Augmentation}

Table~\ref{tab:isolated} presents the main comparison of all isolated
strategies against the Phase~2 baseline. To manage the nine-dataset scope,
we report the macro-averaged CER across PATS-A01 fonts and the KHATT result
separately.

\begin{table}[t]
\caption{Isolated augmentation strategies vs.\ Phase~2 zero-shot baseline.
PATS-A01 CER is the unweighted mean across all eight fonts. KHATT is reported
separately. Diacritics stripped. Best LLM result per column in \textbf{bold}.}
\label{tab:isolated}
\begin{center}
\small
\begin{tabular}{lcccc}
\toprule
& \multicolumn{2}{c}{\textbf{PATS-A01 (avg)}} & \multicolumn{2}{c}{\textbf{KHATT}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
\textbf{Phase} & \textbf{CER} & \textbf{$\Delta$P2} & \textbf{CER} & \textbf{$\Delta$P2} \\
\midrule
1 (OCR baseline)       & 14.2 & ---   & 44.6 & ---  \\
\textbf{2 (Zero-shot)} & 12.3 & ---   & 43.0 & ---  \\
\midrule
3 (OCR-Aware)          & 12.0 & -0.3  & 42.7 & -0.3 \\
\textbf{4 (Self-Reflective)} & \textbf{11.6} & \textbf{-0.7} & 42.8 & -0.2 \\
5 (CAMeL post-proc.)   & 12.4 & +0.1  & 43.1 & +0.1 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\textbf{Phase 3 (OCR-Aware).}
Injecting Qaari's character confusion patterns, filtered to those overlapping
with the model's own training-split failures, yields a modest but consistent
improvement: $-$0.3 pp on both PATS-A01 and KHATT. The effect is larger on
high-error fonts (Naskh, Tahoma) where the confusion patterns are more
frequent and actionable, and smaller on low-error fonts.

\textbf{Phase 4 (Self-Reflective).}
Feeding the model statistical insights about its own failure modes, together
with concrete error-correction pairs from training data, produces the largest
improvement of any isolated strategy on PATS-A01 ($-$0.7 pp, 5.7\% relative).
The self-analysis is computed separately for PATS-A01 and KHATT, yet the KHATT
benefit is modest ($-$0.2 pp). This suggests that KHATT's error diversity
limits the utility of aggregate statistical patterns: the model's failures on
handwritten text are less systematic and harder to summarize into actionable
insights.

\textbf{Phase 5 (CAMeL Post-Processing).}
Morphological validation with known-overcorrection revert produces a small
\textit{degradation} relative to zero-shot ($+$0.1 pp on both datasets).
This follows directly from the error taxonomy: 71\% of OCR errors are
valid-but-wrong, meaning the original OCR tokens pass morphological analysis.
The revert strategy can only intervene when the LLM output is morphologically
invalid while the OCR token is valid---a condition that triggers on only a
minority of error cases, while occasionally reverting beneficial corrections.

\subsection{Phase 6: Combinations and Ablation}

Table~\ref{tab:combos} presents the results of combining multiple knowledge
sources.

\begin{table}[t]
\caption{Phase 6: Strategy combinations vs.\ Phase~2 zero-shot baseline.
conf=confusion-matrix injection, self=self-reflective, camel=CAMeL
post-processing. PATS-A01 CER is the unweighted macro-average. Diacritics
stripped. Best result per column in \textbf{bold}.}
\label{tab:combos}
\begin{center}
\small
\begin{tabular}{lcccc}
\toprule
& \multicolumn{2}{c}{\textbf{PATS-A01 (avg)}} & \multicolumn{2}{c}{\textbf{KHATT}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
\textbf{Combination} & \textbf{CER} & \textbf{$\Delta$P2} & \textbf{CER} & \textbf{$\Delta$P2} \\
\midrule
P2 (zero-shot)   & 12.3 & ---   & 43.0 & ---  \\
conf\_only (P3)   & 12.0 & -0.3  & 42.7 & -0.3 \\
self\_only (P4)   & 11.6 & -0.7  & 42.8 & -0.2 \\
\textbf{conf\_self (P3+P4)}  & \textbf{11.2} & \textbf{-1.1} & \textbf{42.4} & \textbf{-0.6} \\
best\_camel (P3+P4+P5) & 11.1 & -1.2 & 42.5 & -0.5 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The \textbf{conf\_self} combination (Phase~3 + Phase~4) achieves the best
CER on both datasets: 11.2\% on PATS-A01 (a 21.1\% relative improvement over
the OCR baseline, and 8.9\% over zero-shot) and 42.4\% on KHATT (5.0\%
relative improvement over OCR baseline). The two strategies are complementary:
confusion-matrix injection targets known character-pair errors, while
self-reflective prompting targets the model's own systematic correction
failures.

Adding CAMeL post-processing on top of conf\_self (the \textbf{best\_camel}
combination) yields a further marginal gain on PATS-A01 (11.1\%) but a slight
reversal on KHATT (42.5\% vs 42.4\%), confirming that Phase~5 contributes
little once the prompt is already providing strong guidance.

\textbf{Ablation.} Removing self-reflective prompting from conf\_self
degrades PATS-A01 CER by 0.8 pp (from 11.2\% to 12.0\%), establishing it as
the more impactful component. Removing confusion-matrix injection degrades by
0.4 pp (from 11.2\% to 11.6\%). Neither component is redundant; their
combination provides a synergistic benefit.

\subsection{Per-Font Analysis}

To understand font-dependent behavior, Table~\ref{tab:perfont} presents the
best-performing strategy per font alongside the zero-shot and OCR baselines.

\begin{table}[t]
\caption{Best LLM strategy per font (CER, diacritics stripped). $\Delta$ is
vs.\ OCR baseline in percentage points. The optimal strategy varies across
fonts, confirming that no single approach dominates universally.}
\label{tab:perfont}
\begin{center}
\small
\begin{tabular}{lrrrr}
\toprule
\textbf{Font} & \textbf{OCR} & \textbf{P2} & \textbf{Best} & \textbf{Best Strategy} \\
\midrule
Akhbar       & 4.3  & 5.2  & 4.9  & P5 (CAMeL only) \\
Simplified   & 7.9  & 7.3  & 6.8  & conf\_self \\
Thuluth      & 12.1 & 11.2 & 10.5 & conf\_self \\
Traditional  & 14.2 & 2.7  & \textbf{2.4}  & P4 (Self-Reflective) \\
Arial        & 14.8 & 14.7 & 14.3 & P4 (Self-Reflective) \\
Andalus      & 15.3 & 15.1 & 14.6 & conf\_self \\
Tahoma       & 20.6 & 20.5 & 19.9 & P4 (Self-Reflective) \\
Naskh        & 24.3 & 21.5 & 20.8 & conf\_self \\
\midrule
KHATT        & 44.6 & 43.0 & 42.4 & conf\_self \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The per-font breakdown reveals two distinct regimes. On Akhbar (4.3\% baseline
CER), \textit{no LLM strategy improves over raw OCR}: Phase~5 (CAMeL-only,
the most conservative strategy) achieves 4.9\% CER, still worse than the 4.3\%
OCR baseline. This is the clearest example of the over-correction threshold
phenomenon. On Traditional (14.2\% baseline CER), self-reflective prompting
achieves 2.4\% CER---an 83.1\% relative reduction---the single largest
improvement observed in this study.

\subsection{Summary}

The results paint a picture where correction effectiveness depends on the
interaction between font characteristics, baseline error rates, and prompt
strategy. The key patterns are:

\begin{enumerate}
\item Zero-shot correction helps most when baseline CER is moderate to high
      (roughly 10--25\%), where the LLM can identify and fix errors without
      excessive intervention.
\item On low-error fonts (Akhbar, $<$5\% CER), no LLM strategy improves over
      raw OCR; the model introduces more errors than it fixes.
\item On very high-error handwritten input (KHATT, 44.6\% CER), the errors
      are too diverse and structurally different for text-only LLM correction
      to resolve beyond marginal improvements.
\item Self-reflective prompting is the most impactful isolated strategy on
      PATS-A01, while the combination of confusion-matrix injection and
      self-reflective prompting is consistently best across both dataset types.
\end{enumerate}

% -------------------------------------------------------------------------
\section{Discussion}
% -------------------------------------------------------------------------

\subsection{The Over-Correction Threshold}

The font-dependent results point to a quantifiable \textit{over-correction
threshold}~$e^*$. Let~$r_f$ denote the model's fix rate (fraction of existing
errors corrected) and~$r_i$ its introduction rate (fraction of correct tokens
incorrectly changed). The expected CER after correction is approximately
$\text{CER}_{\text{out}} \approx e(1 - r_f) + (1-e)\,r_i$, and correction is
net beneficial when $e \cdot r_f > (1-e) \cdot r_i$, yielding a threshold
$e^* = r_i / (r_f + r_i)$.  Empirically, $r_i \approx 0.22$ and
$r_f \approx 0.76$ across PATS-A01, giving $e^* \approx 22\%$ in aggregate;
however, the per-font crossing between Akhbar (4.3\%, harmed) and Simplified
(7.9\%, helped) places the effective threshold at approximately 6--7\% CER,
reflecting a drop in~$r_f$ at low error densities. For PATS-A01 fonts, the
beneficial correction range is approximately 8--25\% CER. Fonts below this
range are harmed by any LLM intervention, while fonts within the range
(Traditional, Naskh, Thuluth) show clear improvement.

\subsection{Why Typewritten and Handwritten Differ}

KHATT's marginal improvement despite its high baseline error rate is striking.
The explanation lies in the nature of the errors: typewritten OCR errors are
predominantly character-level substitutions (dot confusions, similar shapes)
that preserve word structure, making them recoverable from text context alone.
Handwritten OCR errors include character merges, splits, segmentation
failures, and stroke misinterpretation that fundamentally alter word structure.
These errors often produce output that does not resemble any Arabic word,
giving the LLM little to work with.

This finding implies that effective handwritten OCR correction may require
multi-modal approaches that combine the OCR image with the text output, rather
than text-only post-processing.

\subsection{Prompt-Induced Over-Correction}

When augmentation degrades performance, the mechanism is consistent across
strategies: additional context in the system prompt acts as an implicit signal
that the input text is defective, raising the model's \textit{a priori}
intervention probability. A zero-shot prompt with a conservative instruction
(``return as-is if correct'') maintains a low-intervention prior. Confusion
lists and diagnostic notes shift this prior toward active rewriting, effectively
increasing~$r_i$ in the threshold model above. Because $e^* = r_i / (r_f + r_i)$,
a higher~$r_i$ raises the CER at which correction breaks even, making it
harmful on a wider range of inputs.

This analysis explains a subtle empirical observation: Phase~4 (self-reflective)
is the \emph{only} isolated strategy that improves on Akhbar. Unlike Phase~3,
which provides correction targets that encourage intervention, Phase~4's
diagnostic text explicitly quantifies the model's own over-correction rate
and warns against unnecessary rewrites. This directly targets~$r_i$---lowering
it---rather than only providing correction guidance that raises intervention.
The conf\_self synergy arises because the two components target different
parameters: Phase~3 raises~$r_f$ by directing attention to known confusion
pairs, while Phase~4 lowers~$r_i$ through behavioural self-regulation.
Together, they shift $e^*$ downward: beneficial correction extends to
lower-error fonts.

\subsection{Why CAMeL Post-Processing Has Limited Effect}

The morphological revert strategy's limited impact follows directly from the
error taxonomy: OCR errors are predominantly valid-but-wrong (real Arabic words
that are incorrect in context), not non-words. A context-free morphological
check cannot distinguish a correct word from an incorrect one if both are
morphologically valid. The known-overcorrection revert partially compensates
for this limitation, but the overcorrection list is sparse relative to the
diversity of LLM errors, limiting its coverage. More powerful
approaches---context-aware disambiguation or semantic validation---might
activate on more cases but would also risk introducing their own errors.

\subsection{Practical Implications}

These results suggest a practical deployment strategy: rather than applying
a single LLM correction pipeline uniformly, systems should estimate the
baseline OCR quality and select the correction strategy accordingly.
High-quality OCR output (CER $<$8\%) should be left unchanged or corrected
only with the most conservative strategy. Moderate-to-high-error typewritten
output benefits from the conf\_self combination. High-error handwritten output
may require fine-tuned models or multi-modal correction rather than prompt
engineering alone.

\subsection{Limitations}

This study tests one LLM (Qwen3-4B) and one OCR engine (Qaari). Larger
models may exhibit different over-correction patterns. The PATS-A01 dataset
uses synthetic document images; real-world typewritten documents may introduce
additional noise (scanning artifacts, paper degradation) that changes the
error profile. The evaluation uses CER and WER as proxies for downstream
task performance; domain-specific metrics may tell a different story. Finally,
the self-reflective insights are derived from the same model's training-split
behavior, which may limit generalization to different models or domains.

% -------------------------------------------------------------------------
\section{Conclusion}
% -------------------------------------------------------------------------

We presented a controlled evaluation of five prompt-engineering strategies for
Arabic post-OCR correction across nine datasets spanning an order-of-magnitude
range of baseline CER (4.3--44.6\%). Three findings emerge with implications
beyond the Arabic OCR domain.

First, \textbf{correction effectiveness is governed by a font-dependent
over-correction threshold.} Below approximately 6--7\% CER, the model's
introduction rate on correct tokens exceeds its fix rate on erroneous ones,
producing net degradation. Above this threshold, zero-shot correction yields
substantial improvements---up to 81\% relative CER reduction on Traditional
font. This threshold is a quantifiable property of the model--engine pair,
formally expressible as $e^* = r_i / (r_f + r_i)$, and its existence implies
that LLM correction should be deployed \emph{conditionally}, not universally.

Second, \textbf{knowledge augmentation operates through two distinct
mechanisms}---raising the fix rate (confusion-matrix injection targets known
error patterns) and lowering the introduction rate (self-reflective prompting
warns against over-correction). These mechanisms are orthogonal: ablation
confirms near-additive contributions, and only the self-reflective component
improves the lowest-error font. The best combination (\texttt{conf\_self})
achieves 11.2\% CER on PATS-A01 ($-$21.1\% relative vs.\ OCR baseline) and
42.4\% on KHATT ($-$5.0\% relative), with statistical significance on six of
nine datasets at Bonferroni-corrected $\alpha$.

Third, \textbf{text-only correction is structurally limited on handwritten
input.} KHATT's 37\% segmentation error rate---word merges and splits that
destroy word boundaries---forms an irreducible error floor for any text-only
corrector, explaining why the highest-CER dataset yields the lowest relative
improvement.

The font-dependent behaviour we observe---invisible in single-dataset
studies---underscores the necessity of multi-font, multi-script evaluation for
post-OCR correction research. Future work should characterise the
over-correction threshold across model scales and OCR engines, explore
confidence-guided selective correction that restricts intervention to
low-confidence tokens, and investigate multi-modal correction that combines
document images with OCR text for handwritten text recovery.

% -------------------------------------------------------------------------
\section*{Acknowledgment}
% -------------------------------------------------------------------------
% Add funding/acknowledgment here when known.

% -------------------------------------------------------------------------
\begin{thebibliography}{00}

\bibitem{qaari}
Qaari OCR Engine, open-source Arabic OCR system.
Available: \url{https://github.com/quran/qaari}

\bibitem{arabic_ocr_survey}
A. Lorigo and V. Govindaraju, ``Offline Arabic handwriting recognition: a survey,''
\textit{IEEE Trans. Pattern Anal. Mach. Intell.}, vol. 28, no. 5, pp. 712--724, 2006.

\bibitem{pats}
M. Waly, A. Atwan, M. Cheriet, and O. Muda, ``PATS-A01: a large-scale Arabic
multi-font printed text corpus,'' in \textit{Proc. Int. Conf. Document Analysis
and Recognition (ICDAR)}, 2019.

\bibitem{khatt}
S. Mahmoud, I. Ahmad, W. Al-Khatib, M. Alshayeb, M. Parvez,
V. Margner, and G. Fink, ``KHATT: an open Arabic offline handwriting dataset,''
\textit{Pattern Recognition}, vol. 47, no. 3, pp. 1096--1112, 2014.

\bibitem{obeid2020}
O. Obeid, N. Zalmout, S. Khalifa, S. Taji, N. Habash, A. Pasha, R. Roth,
M. Diab, and K. Choukri, ``CAMeL tools: an open source python toolkit for
Arabic natural language processing,'' in \textit{Proc. LREC}, 2020, pp. 7022--7032.

\bibitem{qwen3}
Qwen Team, ``Qwen3 technical report,'' \textit{arXiv preprint arXiv:2505.09388},
2025.

\bibitem{afli2016}
H. Afli, L. Barrault, and H. Schwenk, ``OCR error correction using statistical
machine translation,'' \textit{Int. J. Comput. Linguistics Appl.}, vol. 7, no. 1,
pp. 175--191, 2016.

\bibitem{dong2018}
R. Dong, O. Lal, and G. Prud'hommeaux, ``Adapting sequence to sequence models
for text normalization in speech recognition,'' in \textit{Proc. INTERSPEECH}, 2018.

\bibitem{nguyen2021}
T. Nguyen, A. Jatowt, M. Coustaty, and A. Doucet, ``Survey of post-OCR processing
approaches,'' \textit{ACM Comput. Surv.}, vol. 54, no. 6, pp. 1--37, 2021.

\bibitem{llm_ocr_correction}
N. Kang, E. Appel, A. Bhatt, J. Kim, S. Gulati, M. Krallinger, and
E. Kim, ``Large language models are better at OCR post-correction than you
might think,'' in \textit{Proc. COLING}, 2024.

\bibitem{arabic_bert_ocr}
W. Antoun, F. Baly, and H. Hajj, ``AraBERT: transformer-based model for
Arabic language understanding,'' in \textit{Proc. LREC Workshop on Arabic
Natural Language Processing}, 2020.

\bibitem{brown2020}
T. Brown \textit{et al.}, ``Language models are few-shot learners,''
\textit{Advances in Neural Information Processing Systems}, vol. 33, 2020.

\bibitem{wei2022}
J. Wei \textit{et al.}, ``Chain-of-thought prompting elicits reasoning in large
language models,'' \textit{Advances in Neural Information Processing Systems},
vol. 35, 2022.

\bibitem{lewis2020}
P. Lewis \textit{et al.}, ``Retrieval-augmented generation for knowledge-intensive
NLP tasks,'' \textit{Advances in Neural Information Processing Systems}, vol. 33,
2020.

\bibitem{shinn2023}
N. Shinn, F. Cassano, B. Labash, A. Gopinath, K. Narasimhan, and
S. Yao, ``Reflexion: language agents with verbal reinforcement learning,''
\textit{Advances in Neural Information Processing Systems}, vol. 36, 2023.

\bibitem{habash2010}
N. Habash, ``Introduction to Arabic natural language processing,''
\textit{Synthesis Lectures on Human Language Technologies}, vol. 3, no. 1,
pp. 1--187, 2010.

\end{thebibliography}

\end{document}
